---
title: "ODOT Safety Analysis"
format: html
editor: visual
---

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
library(sf)
library(leaflet)
library(tidyverse)
setwd("~/2025 - 2026 XU/project")

### Read Road_inventory

road_inventory <- st_read("Road_Inventory.gdb", layer = "RoadInventory")

### Reading and combining Crash Data

Crash2020 <- read.csv("crashes/2020Crash.csv")
Crash2021 <- read.csv("crashes/2021Crash.csv")
Crash2022 <- read.csv("crashes/2022Crash.csv")
Crash2023 <- read.csv("crashes/2023Crash.csv")
Crash2024 <- read.csv("crashes/2024Crash.csv")

crashes_all <- rbind(
  Crash2020,
  Crash2021,
  Crash2022,
  Crash2023,
  Crash2024
)

### Read in Projects

All_Projects <- read.csv("All_Projects.csv")
head(All_Projects)

### Reading in PCR

st_layers("PCR_Local.gdb")
PCR_Local <- st_read("PCR_Local.gdb", layer = "PCR_Local")

st_layers("PCR_State.gdb")
PCR_State <- st_read("PCR_State.gdb", layer = "PCRState")

# Original ID columns list
id_cols <- c("PCR_HISTORY_ID", "ROUTE_NBR", "ROUTE_SUFFIX", "NLFID", "COUNTY_CD",
             "JURISDICTION", "ROUTE_TYPE", "DIRECTION_CD", "LEFT_MUNI_NAME", 
             "CITY_NAME", "LEFT_TWP_NAME", "TOWNSHIP_NAME")

# Keep only columns that exist in the dataframe
id_cols_exist <- intersect(id_cols, colnames(PCR_Local))

# Function to safely convert character to numeric except ID columns
make_numeric_safe <- function(df, exclude_cols = id_cols_exist) {
  df %>%
    mutate(across(.cols = where(is.character) & !all_of(exclude_cols), 
                  ~ suppressWarnings(as.numeric(.))))
}

# Apply to both datasets
PCR_Local_fixed <- make_numeric_safe(PCR_Local)
PCR_State_fixed <- make_numeric_safe(PCR_State)

PCR_combined <- bind_rows(
  PCR_Local_fixed %>% mutate(Source = "Local"),
  PCR_State_fixed %>% mutate(Source = "State")
)

#write.csv(crashes_all, "crashes_all.csv", row.names = FALSE)
#write.csv(PCR_combined, "PCR_combined.csv", row.names = FALSE)
#write.csv(road_inventory, "road_inventory.csv", row.names = FALSE)

head(road_inventory)


############################################
### BIG JOIN: CRASHES + ROAD INVENTORY + PCR
############################################

### ----------------------------------------
### 1. LOAD CORE DATASETS
### ----------------------------------------

# Crashes
crashes <- crashes_all

# PCR
pcr <- PCR_combined

### ----------------------------------------
### 2. CONVERT CRASHES TO SPATIAL FEATURES
### ----------------------------------------

# Convert crash table to sf points using lat/long
crashes_sf <- st_as_sf(
  crashes,
  coords = c("ODOT_LONGITUDE_NBR", "ODOT_LATITUDE_NBR"),
  crs = 4326,
  remove = FALSE
)

# Check CRS consistency
st_crs(crashes_sf)
st_crs(road_inventory)

# Project crashes to match road inventory CRS (EPSG:3857)
crashes_sf <- st_transform(crashes_sf, 3857)

# Drop Z/M dimensions (required for GEOS spatial joins)
crashes_sf <- st_zm(crashes_sf, drop = TRUE, what = "ZM")
road_inventory <- st_zm(road_inventory, drop = TRUE, what = "ZM")

### ----------------------------------------
### 3. SPATIAL JOIN: CRASHES → ROAD SEGMENTS
### ----------------------------------------

# Match each crash to its nearest road inventory segment
crash_segment <- st_join(
  crashes_sf,
  road_inventory,
  join = st_nearest_feature
)

# Quick sanity check
head(crash_segment)

# Optional export
#write.csv(crash_segment, "crash_segment.csv", row.names = FALSE)

### ----------------------------------------
### 4. AGGREGATE CRASHES TO SEGMENT LEVEL
### ----------------------------------------

# Collapse crash-level data into one row per road segment
segment_crash_summary <- crash_segment %>%
  st_drop_geometry() %>%
  group_by(ROADWAY_INVENTORY_ID) %>%
  summarise(
    total_crashes = n(),
    fatal_crashes = sum(CRASH_SEVERITY_CD == 1, na.rm = TRUE),
    serious_injury_crashes = sum(CRASH_SEVERITY_CD == 2, na.rm = TRUE),
    injury_crashes = sum(CRASH_SEVERITY_CD %in% 1:4, na.rm = TRUE)
  )

# Optional export
# write.csv(crash_segment, "crash_segment.csv", row.names = FALSE)


### ----------------------------------------
### 5. JOIN CRASH COUNTS BACK TO ROAD NETWORK
### ----------------------------------------
road_with_crashes <- road_inventory %>%
  left_join(
    segment_crash_summary,
    by = "ROADWAY_INVENTORY_ID"
  )

road_with_crashes <- road_with_crashes %>%
  mutate(
    total_crashes = replace_na(total_crashes, 0),
    fatal_crashes = replace_na(fatal_crashes, 0),
    serious_injury_crashes = replace_na(serious_injury_crashes, 0),
    injury_crashes = replace_na(injury_crashes, 0)
  )


### ----------------------------------------
### 6. PREPARE PCR DATA (LATEST PER SEGMENT)
### ----------------------------------------

# Drop geometry for attribute join
PCR_df <- st_drop_geometry(PCR_combined)
class(PCR_df)

# Align key name with road inventory
PCR_df <- PCR_df %>%
  rename(NLF_ID = NLFID)

# Keep only the most recent PCR observation per segment
PCR_latest <- PCR_df %>%
  group_by(NLF_ID) %>%
  slice_max(order_by = PCR_YEAR, n = 1, with_ties = FALSE) %>%
  ungroup()


### ----------------------------------------
### 7. JOIN PCR DATA TO ROAD + CRASH DATA
### ----------------------------------------

final_sf <- road_with_crashes %>%
  left_join(
    PCR_latest,
    by = "NLF_ID"
  )

# Verify row count preserved (1 row per segment)
nrow(final_sf) == nrow(road_with_crashes)


### ----------------------------------------
### 8. COMPUTE CRASH RATES
### ----------------------------------------

# Crashes per mile
final_sf <- final_sf %>%
  mutate(
    crashes_per_mile = total_crashes / SEGMENT_LENGTH_NBR
  )

# Million Vehicle Miles Traveled (MVMT) and crash rate
final_sf <- final_sf %>%
  mutate(
    MVMT = (AADT_TOTAL * 365 * SEGMENT_LENGTH_NBR) / 1e6,
    crashes_per_MVMT = total_crashes / MVMT
  )


### ----------------------------------------
### 9. SANITY CHECKS
### ----------------------------------------

summary(final_sf$total_crashes)
summary(final_sf$crashes_per_mile)
summary(final_sf$crashes_per_MVMT)

sum(is.infinite(final_sf$crashes_per_MVMT))

segment_crash_summary <- crash_segment %>%
  st_drop_geometry() %>%
  group_by(ROADWAY_INVENTORY_ID) %>%
  summarise(
    total_crashes = n(),
    fatal_crashes = sum(CRASH_SEVERITY_TXT == "Fatal", na.rm = TRUE),
    injury_crashes = sum(CRASH_SEVERITY_CD %in% c(2,3,4), na.rm = TRUE),
    .groups = "drop"
  )

road_with_crashes <- road_inventory %>%
  left_join(segment_crash_summary, by = "ROADWAY_INVENTORY_ID") %>%
  mutate(
    total_crashes = replace_na(total_crashes, 0),
    fatal_crashes = replace_na(fatal_crashes, 0),
    injury_crashes = replace_na(injury_crashes, 0)
  )

road_with_crashes <- road_with_crashes %>%
  mutate(
    crashes_per_mile = total_crashes / SEGMENT_LENGTH_NBR,
    crashes_per_MVMT = (total_crashes * 1e6) /
      (ADT_TOTAL_NBR * 365 * SEGMENT_LENGTH_NBR)
  )


### ----------------------------------------
### 10. EXPORT FINAL DATASET
### ----------------------------------------

final_sf <- road_with_crashes %>%
  left_join(PCR_latest, by = "NLF_ID")


#Sanity Checks
summary(road_with_crashes$crashes_per_mile)
summary(road_with_crashes$crashes_per_MVMT)

sum(is.na(road_with_crashes$crashes_per_MVMT))
sum(is.infinite(road_with_crashes$crashes_per_MVMT))

##write.csv(final_sf, "final_sf", row.names = FALSE)

##write.csv(
  #final_sf[1:25000, ],
  #"final_sf_first_25k.csv",
 # row.names = FALSE
#)

############################################
### ANALYSIS-READY SUBSET
############################################

analysis_df <- final_sf %>%
  filter(
    !is.na(ADT_TOTAL_NBR),
    ADT_TOTAL_NBR > 0,
    SEGMENT_LENGTH_NBR > 0
  )


library(lubridate)

projects_filtered <- All_Projects %>%
  mutate(
    END_CONSTR_MILESTONE_DT = as.Date(END_CONSTR_MILESTONE_DT)
  ) %>%
  filter(
    !is.na(END_CONSTR_MILESTONE_DT),
    year(END_CONSTR_MILESTONE_DT) >= 2021,
    year(END_CONSTR_MILESTONE_DT) <= 2023
  )

nrow(projects_filtered)

temp_join <- analysis_df %>%
  inner_join(projects_filtered, by = "NLF_ID")

names(temp_join)

treated_segments <- temp_join %>%
  filter(
    CTL_BEGIN_NBR <= CTL_END.y,
    CTL_END_NBR >= CTL_BEGIN.y
  )

nrow(treated_segments)

treated_ids <- treated_segments %>%
  distinct(ROADWAY_INVENTORY_ID)

nrow(treated_ids)

analysis_df <- analysis_df %>%
  mutate(
    treated = ROADWAY_INVENTORY_ID %in% treated_ids$ROADWAY_INVENTORY_ID
  )
table(analysis_df$treated)

summary(as.Date(treated_segments$END_CONSTR_MILESTONE_DT))
```

```{r}
############################################
### yearly crash rates per segment
############################################

segment_year_crashes <- crash_segment %>%
  st_drop_geometry() %>%
  group_by(ROADWAY_INVENTORY_ID, CRASH_YR) %>%
  summarise(
    yearly_crashes = n(),
    yearly_fatal = sum(CRASH_SEVERITY_CD == 1, na.rm = TRUE),
    yearly_injury = sum(CRASH_SEVERITY_CD %in% c(2,3,4), na.rm = TRUE),
    .groups = "drop"
  )

head(segment_year_crashes)

segment_year_data <- segment_year_crashes %>%
  left_join(
    analysis_df %>%
      st_drop_geometry() %>%
      select(ROADWAY_INVENTORY_ID,
             SEGMENT_LENGTH_NBR,
             ADT_TOTAL_NBR),
    by = "ROADWAY_INVENTORY_ID"
  )

segment_year_data <- segment_year_data %>%
  mutate(
    MVMT = (ADT_TOTAL_NBR * 365 * SEGMENT_LENGTH_NBR) / 1e6,
    crash_rate_MVMT = yearly_crashes / MVMT
  )

summary(segment_year_data$crash_rate_MVMT)

segment_year_data <- segment_year_data %>%
  mutate(
    period = case_when(
      CRASH_YR %in% c(2020, 2021) ~ "before",
      CRASH_YR %in% c(2023, 2024) ~ "after",
      TRUE ~ "transition"
    )
  )

segment_year_data <- segment_year_data %>%
  filter(period != "transition")

table(segment_year_data$CRASH_YR)

treated_ids <- treated_segments %>%
  distinct(ROADWAY_INVENTORY_ID)

segment_year_data <- segment_year_data %>%
  mutate(
    treated = ROADWAY_INVENTORY_ID %in% treated_ids$ROADWAY_INVENTORY_ID
  )

table(segment_year_data$treated)

did_summary <- segment_year_data %>%
  group_by(treated, period) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    median_crash_rate = median(crash_rate_MVMT, na.rm = TRUE),
    observations = n(),
    .groups = "drop"
  )

did_summary

did_effect <- did_summary %>%
  select(treated, period, avg_crash_rate) %>%
  tidyr::pivot_wider(
    names_from = period,
    values_from = avg_crash_rate
  ) %>%
  mutate(
    change = after - before
  )

did_effect

### did_effect: 1.37 based on this sample of 2021–2023 projects, treated 
## segments did not show greater crash reduction than untreated segments.

```

You can add options to executable code like this

```{r}
############################################
### Projects using Safety Funds
############################################



projects_safety <- projects_filtered %>%
  filter(SAFETY_FUNDS_IND == "Y")

nrow(projects_safety)

treated_safety_segments <- analysis_df %>%
  inner_join(projects_safety, by = "NLF_ID") %>%
  filter(
    CTL_BEGIN_NBR <= CTL_END.y,
    CTL_END_NBR >= CTL_BEGIN.y
  )

treated_safety_ids <- treated_safety_segments %>%
  distinct(ROADWAY_INVENTORY_ID)

segment_year_data <- segment_year_data %>%
  mutate(
    treated_safety =
      ROADWAY_INVENTORY_ID %in%
      treated_safety_ids$ROADWAY_INVENTORY_ID
  )

table(segment_year_data$treated_safety)

did_summary_safety <- segment_year_data %>%
  group_by(treated_safety, period) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    median_crash_rate = median(crash_rate_MVMT, na.rm = TRUE),
    observations = n(),
    .groups = "drop"
  )

did_effect_safety <- did_summary_safety %>%
  select(treated_safety, period, avg_crash_rate) %>%
  tidyr::pivot_wider(
    names_from = period,
    values_from = avg_crash_rate
  ) %>%
  mutate(
    change = after - before
  )

did_safety_effect_value <-
  did_effect_safety$change[did_effect_safety$treated_safety == TRUE] -
  did_effect_safety$change[did_effect_safety$treated_safety == FALSE]
did_safety_effect_value

### also suggests that construction isnt helping and making it worse
### raises the question do safety projects show immediate improvements, or is there a lag effect
```

```{r}
############################################
### Crash Rate Around Safety Project Completion
############################################

projects_safety <- projects_safety %>%
  mutate(
    completion_year = year(as.Date(END_CONSTR_MILESTONE_DT))
  ) %>%
  filter(!is.na(completion_year))

treated_safety_segments <- analysis_df %>%
  inner_join(projects_safety, by = "NLF_ID") %>%
  filter(
    CTL_BEGIN_NBR <= CTL_END.y,
    CTL_END_NBR >= CTL_BEGIN.y
  ) %>%
  select(ROADWAY_INVENTORY_ID, completion_year) %>%
  distinct()

segment_year_data <- segment_year_data %>%
  left_join(
    treated_safety_segments,
    by = "ROADWAY_INVENTORY_ID"
  )

segment_year_data <- segment_year_data %>%
  mutate(
    event_time = CRASH_YR - completion_year
  )

event_study_data <- segment_year_data %>%
  filter(
    !is.na(event_time),
    event_time >= -2,
    event_time <= 2
  )

event_summary <- event_study_data %>%
  group_by(event_time) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    observations = n(),
    .groups = "drop"
  )

ggplot(event_summary, aes(x = event_time, y = avg_crash_rate)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Crash Rate Around Safety Project Completion",
    x = "Years Relative to Project Completion",
    y = "Average Crash Rate (per MVMT)"
  ) +
  theme_minimal()
```

```{r}
############################################
### Event Study: Treated vs Control
############################################

# segment_year_data already has completion_year and event_time from above
# DO NOT re-join treated_safety_segments here — it was already joined

event_window <- segment_year_data %>%
  filter(
    treated_safety == TRUE,
    event_time >= -2,
    event_time <= 2
  )

# For control, anchor their "event time" to the median completion year of treated projects
median_completion_year <- median(
  treated_safety_segments$completion_year, na.rm = TRUE
)

control_window <- segment_year_data %>%
  filter(treated_safety == FALSE) %>%
  mutate(
    event_time = CRASH_YR - median_completion_year  # <-- THIS was the bug
  ) %>%
  filter(event_time >= -2, event_time <= 2)

event_study_df <- bind_rows(
  event_window %>% mutate(group = "Treated (Safety)"),
  control_window %>% mutate(group = "Control")
)

event_summary <- event_study_df %>%
  group_by(group, event_time) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )

ggplot(event_summary, aes(x = event_time, y = avg_crash_rate, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous(breaks = -2:2) +
  scale_y_continuous(
    limits = c(0, 20),
    breaks = 5:18
  ) +
  labs(
    title = "Event Study: Crash Rate Around Safety Project Completion",
    subtitle = "t = 0 is the year construction completed",
    x = "Years Relative to Project Completion",
    y = "Average Crash Rate (per MVMT)",
    color = "Group"
  ) +
  theme_minimal()

library(lfe)

did_model <- felm(
  crash_rate_MVMT ~ treated_safety * I(period == "after") |
    ROADWAY_INVENTORY_ID + CRASH_YR |
    0 |
    ROADWAY_INVENTORY_ID,
  data = segment_year_data
)
summary(did_model)

#Safety-funded segments showed essentially zero additional change in crash rate compared to control segments
#The effect is statistically indistinguishable from zero
#This confirms your raw DiD finding

# Keep only segments observed in BOTH before and after periods
balanced_segments <- segment_year_data %>%
  group_by(ROADWAY_INVENTORY_ID) %>%
  filter(all(c("before", "after") %in% period)) %>%
  ungroup()

nrow(balanced_segments)  # check how many remain

did_model_balanced <- felm(
  crash_rate_MVMT ~ treated_safety * I(period == "after") |
    ROADWAY_INVENTORY_ID + CRASH_YR |
    0 |
    ROADWAY_INVENTORY_ID,
  data = balanced_segments
)

summary(did_model_balanced)

### no noticable changes

# Clean event study - treated segments only, no re-joining
# Clean event study - treated segments only, no re-joining
event_summary <- segment_year_data %>%
  filter(
    treated_safety == TRUE,
    !is.na(event_time),
    event_time >= -2,
    event_time <= 2
  ) %>%
  group_by(event_time) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    se = sd(crash_rate_MVMT, na.rm = TRUE) / sqrt(n()),
    n_obs = n(),
    .groups = "drop"
  )

print(event_summary)

ggplot(event_summary, aes(x = event_time, y = avg_crash_rate)) +
  geom_ribbon(aes(ymin = avg_crash_rate - 1.96*se,
                  ymax = avg_crash_rate + 1.96*se),
              alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.1, y = max(event_summary$avg_crash_rate),
           label = "Construction\nComplete", hjust = 0, color = "red", size = 5.5) +
  scale_x_continuous(breaks = -2:2) +
  labs(
    title = "Event Study: Crash Rates Around Safety Project Completion",
    subtitle = "Treated segments only | t=0 is completion year",
    x = "Years Relative to Project Completion",
    y = "Average Crash Rate (per MVMT)"
  ) +
  theme_minimal()


# Filter to only segments where construction completed in 2021
# This gives us 3 full years of post-completion data (2022, 2023, 2024)

# Filter to only segments where construction completed in 2021
# This gives us 3 full years of post-completion data (2022, 2023, 2024)

cohort_2021 <- segment_year_data %>%
  filter(
    treated_safety == TRUE,
    !is.na(completion_year),
    completion_year == 2021
  ) %>%
  mutate(event_time = CRASH_YR - 2021)

# Check what years we have
table(cohort_2021$CRASH_YR)
print(paste("Number of segments:", n_distinct(cohort_2021$ROADWAY_INVENTORY_ID)))

cohort_summary <- cohort_2021 %>%
  filter(event_time >= -2, event_time <= 3) %>%  # extend to +3 for lag testing
  group_by(event_time) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    se = sd(crash_rate_MVMT, na.rm = TRUE) / sqrt(n()),
    n_obs = n(),
    .groups = "drop"
  )

print(cohort_summary)

ggplot(cohort_summary, aes(x = event_time, y = avg_crash_rate)) +
  geom_ribbon(aes(ymin = avg_crash_rate - 1.96*se,
                  ymax = avg_crash_rate + 1.96*se),
              alpha = 0.2, fill = "darkgreen") +
  geom_line(linewidth = 1.2, color = "darkgreen") +
  geom_point(size = 3, color = "darkgreen") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.1, y = max(cohort_summary$avg_crash_rate),
           label = "Construction\nComplete", hjust = 0, color = "red", size = 5.5) +
  scale_x_continuous(breaks = -2:3) +
  labs(
    title = "2021 Cohort: Crash Rates After Safety Project Completion",
    subtitle = "Only segments completing construction in 2021 | Extends to t=+3",
    x = "Years Relative to Project Completion (2021)",
    y = "Average Crash Rate (per MVMT)"
  ) +
  theme_minimal()



# Compare baseline characteristics of treated vs control
baseline_check <- analysis_df %>%
  st_drop_geometry() %>%
  mutate(
    treated_safety = ROADWAY_INVENTORY_ID %in% treated_safety_ids$ROADWAY_INVENTORY_ID
  ) %>%
  group_by(treated_safety) %>%
  summarise(
    n_segments = n(),
    avg_ADT = mean(ADT_TOTAL_NBR, na.rm = TRUE),
    avg_length = mean(SEGMENT_LENGTH_NBR, na.rm = TRUE),
    avg_crash_rate = mean(crashes_per_MVMT, na.rm = TRUE),
    avg_PCR = mean(PCR_NBR, na.rm = TRUE),
    .groups = "drop"
  )

print(baseline_check)

### Compared roads had bias so we need to reapproach

# Step 1: Get the ADT range of treated segments
treated_adt_range <- analysis_df %>%
  st_drop_geometry() %>%
  mutate(
    treated_safety = ROADWAY_INVENTORY_ID %in% treated_safety_ids$ROADWAY_INVENTORY_ID
  ) %>%
  filter(treated_safety == TRUE) %>%
  summarise(
    low  = quantile(ADT_TOTAL_NBR, 0.10, na.rm = TRUE),
    high = quantile(ADT_TOTAL_NBR, 0.90, na.rm = TRUE)
  )

print(treated_adt_range)
# Expected output: low = 573, high = 39364

# Step 2: Filter to comparable roads only
matched_data <- segment_year_data %>%
  filter(
    treated_safety == TRUE |
    (treated_safety == FALSE &
     ADT_TOTAL_NBR >= treated_adt_range$low &
     ADT_TOTAL_NBR <= treated_adt_range$high)
  )

table(matched_data$treated_safety)
# Expected: FALSE ~155k, TRUE ~14k

# Step 3: DiD summary on matched sample
did_summary_matched <- matched_data %>%
  group_by(treated_safety, period) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )

matched_data <- segment_year_data %>%
  filter(
    treated_safety == TRUE |
      (treated_safety == FALSE &
         ADT_TOTAL_NBR >= treated_adt_range$low &
         ADT_TOTAL_NBR <= treated_adt_range$high)
  )

table(matched_data$treated_safety)

did_summary_matched <- matched_data %>%
  group_by(treated_safety, period) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )

print(did_summary_matched)

did_plot_data <- data.frame(
  period = c("Before", "After", "Before", "After"),
  group = c("Control", "Control", "Treated (Safety-Funded)", "Treated (Safety-Funded)"),
  avg_crash_rate = c(10.6, 10.3, 7.65, 7.42)
) %>%
  mutate(period = factor(period, levels = c("Before", "After")))

ggplot(did_plot_data, aes(x = period, y = avg_crash_rate, 
                          color = group, group = group)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 4) +
  geom_label(aes(label = round(avg_crash_rate, 2)), 
             nudge_x = 0.08, show.legend = FALSE, size = 5.5) +
  scale_color_manual(values = c("Control" = "gray50", 
                                "Treated (Safety-Funded)" = "steelblue")) +
  labs(
    title = "Difference-in-Differences: Safety Project Impact on Crash Rates",
    subtitle = "Matched on ADT volume | DiD effect = +0.07 (not significant)",
    x = NULL,
    y = "Average Crash Rate (per MVMT)",
    color = "Group"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

```{r}
#################################
### Looking at PRIMARY_WORK_TYPE
#################################

# Step 1: Create projects_safety_typed
projects_safety_typed <- projects_safety %>%
  select(NLF_ID, CTL_BEGIN, CTL_END, completion_year, PRIMARY_WORK_CATEGORY)

# Step 1: do the join first and check what columns exist
test_join <- analysis_df %>%
  st_drop_geometry() %>%
  inner_join(projects_safety_typed, by = "NLF_ID", relationship = "many-to-many")

# Check if CTL_BEGIN/CTL_END survived the join
names(test_join)[grep("CTL", names(test_join))]

treated_typed_segments <- analysis_df %>%
  st_drop_geometry() %>%
  inner_join(projects_safety_typed, by = "NLF_ID", relationship = "many-to-many") %>%
  filter(
    CTL_BEGIN_NBR <= CTL_END.y,
    CTL_END_NBR >= CTL_BEGIN.y
  ) %>%
  select(ROADWAY_INVENTORY_ID, completion_year, PRIMARY_WORK_CATEGORY) %>%
  distinct(ROADWAY_INVENTORY_ID, PRIMARY_WORK_CATEGORY, .keep_all = TRUE)

# Check project type counts
category_counts <- treated_typed_segments %>%
  count(PRIMARY_WORK_CATEGORY, sort = TRUE)

View(category_counts)

# Categories with enough data to analyze
key_categories <- c(
  "Traffic Control (Safety)",
  "Roadway Minor Rehab", 
  "Intersection Improvement (Safety)",
  "Roadside / Median Improvement (Safety)",
  "Pedestrian Facilities",
  "Pavement Treatments (Safety)"
)

# Join category info onto segment_year_data
segment_year_typed <- segment_year_data %>%
  left_join(
    treated_typed_segments %>% 
      filter(PRIMARY_WORK_CATEGORY %in% key_categories),
    by = "ROADWAY_INVENTORY_ID"
  ) %>%
  mutate(
    work_category = case_when(
      !is.na(PRIMARY_WORK_CATEGORY) ~ PRIMARY_WORK_CATEGORY,
      treated_safety == FALSE ~ "Control",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(work_category))

# Calculate control change separately first
control_change_val <- segment_year_typed %>%
  filter(work_category == "Control") %>%
  group_by(period) %>%
  summarise(avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = period, values_from = avg_crash_rate) %>%
  mutate(change = after - before) %>%
  pull(change)

cat("Control change:", control_change_val, "\n")

# Now run DiD for each category
did_by_category <- segment_year_typed %>%
  filter(work_category %in% key_categories) %>%
  group_by(work_category, period) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = period, values_from = c(avg_crash_rate, n_obs)) %>%
  mutate(
    change = avg_crash_rate_after - avg_crash_rate_before,
    did_effect = change - control_change_val
  )

as.data.frame(did_by_category)

ggplot(did_by_category, aes(x = reorder(work_category, did_effect), 
                            y = did_effect, 
                            fill = did_effect < 0)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_text(aes(label = round(did_effect, 2)), 
            hjust = ifelse(did_by_category$did_effect < 0, 1.2, -0.2),
            size = 4) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
                    labels = c("FALSE" = "Crash increase", "TRUE" = "Crash reduction")) +
  labs(
    title = "Safety Project Impact on Crash Rates by Work Category",
    subtitle = "DiD effect vs control | Negative = crash reduction relative to control",
    x = NULL,
    y = "DiD Effect (crashes per MVMT vs control)",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

### Very interesting stuff here, basically seems to be showing me that only intersection
### improvement projects are making things safer where as other safety projects make 
### little changes or even make things worse. 
```

```{r}
##################################
### What makes roads Safer? (RF)
##################################

# Build a modeling dataset at the segment level
# Join all the info we have onto segments that have crash rate changes

# First create a before/after crash rate per segment
segment_change <- segment_year_data %>%
  filter(period %in% c("before", "after")) %>%
  group_by(ROADWAY_INVENTORY_ID, period) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = period, values_from = avg_crash_rate) %>%
  mutate(
    crash_rate_change = after - before,
    improved = crash_rate_change < 0  # TRUE = got safer
  ) %>%
  filter(!is.na(before), !is.na(after))

library(randomForest)

# Build clean modeling dataset
model_df <- segment_change %>%
  left_join(
    analysis_df %>%
      st_drop_geometry() %>%
      select(
        ROADWAY_INVENTORY_ID,
        ADT_TOTAL_NBR,
        SEGMENT_LENGTH_NBR,
        FUNCTION_CLASS_CD,
        SPEED_LIMIT_NBR,
        MEDIAN_TYPE_CD,
        LANES_NBR,
        LANE_WIDTH_NBR,
        SHOULDER_TYPE_CD,
        SHOULDER_TOTAL_WIDTH_LT,
        SHOULDER_TOTAL_WIDTH_RT,
        SURFACE_TYPE_LEFT_CD,
        AREA_TYPE,
        DIVIDED_HWY_IND.x,
        PCR_NBR,
        AVG_IRI,
        TOT_PCR_DEDUCT_NBR,
        TOT_CRACK_DEDUCT_NBR,
        TOT_RUT_DEDUCT_NBR,
        RUTTING_VALUE,
        POTHOLE_VALUE,
        PATCHING_VALUE,
        crashes_per_MVMT
      ),
    by = "ROADWAY_INVENTORY_ID"
  ) %>%
  left_join(
    treated_typed_segments %>%
      select(ROADWAY_INVENTORY_ID, PRIMARY_WORK_CATEGORY),
    by = "ROADWAY_INVENTORY_ID"
  ) %>%
  mutate(
    PRIMARY_WORK_CATEGORY = replace_na(PRIMARY_WORK_CATEGORY, "No Project"),
    PRIMARY_WORK_CATEGORY = as.factor(PRIMARY_WORK_CATEGORY),
    FUNCTION_CLASS_CD = as.factor(FUNCTION_CLASS_CD),
    MEDIAN_TYPE_CD = as.factor(MEDIAN_TYPE_CD),
    SHOULDER_TYPE_CD = as.factor(SHOULDER_TYPE_CD),
    SURFACE_TYPE_LEFT_CD = as.factor(SURFACE_TYPE_LEFT_CD),
    AREA_TYPE = as.factor(AREA_TYPE),
    DIVIDED_HWY_IND.x = as.factor(DIVIDED_HWY_IND.x),
    improved = as.factor(ifelse(crash_rate_change < 0, "Improved", "Worsened"))
  )

# Check outcome balance
table(model_df$improved)

# Check missingness
colSums(is.na(model_df %>% select(-ROADWAY_INVENTORY_ID, -before, -after, 
                                  -crash_rate_change)))


# Drop high-NA columns, keep clean ones
rf_df <- model_df %>%
  select(
    improved,
    ADT_TOTAL_NBR,
    SEGMENT_LENGTH_NBR,
    FUNCTION_CLASS_CD,
    SPEED_LIMIT_NBR,
    LANES_NBR,
    SURFACE_TYPE_LEFT_CD,
    DIVIDED_HWY_IND.x,
    PCR_NBR,
    TOT_PCR_DEDUCT_NBR,
    TOT_CRACK_DEDUCT_NBR,
    TOT_RUT_DEDUCT_NBR,
    crashes_per_MVMT,
    PRIMARY_WORK_CATEGORY
  ) %>%
  # Drop rows with any remaining NAs
  drop_na()

# Check what's left
nrow(rf_df)
table(rf_df$improved)

# Train random forest
set.seed(42)
rf_model <- randomForest(
  improved ~ .,
  data = rf_df,
  ntree = 500,
  importance = TRUE,
  sampsize = c(Improved = 3000, Worsened = 3000)  # balance the classes
)

print(rf_model)

# Extract feature importance
importance_df <- as.data.frame(importance(rf_model)) %>%
  rownames_to_column("Variable") %>%
  arrange(desc(MeanDecreaseGini))

# Plot
ggplot(importance_df, aes(x = reorder(Variable, MeanDecreaseGini), 
                          y = MeanDecreaseGini)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "What Predicts Whether a Road Segment Improves or Worsens?",
    subtitle = "Random Forest Feature Importance (Mean Decrease Gini)",
    x = NULL,
    y = "Importance Score"
  ) +
  theme_minimal(base_size = 12)

# Compare characteristics between improved and worsened segments directly
rf_df %>%
  group_by(improved) %>%
  summarise(
    avg_ADT = mean(ADT_TOTAL_NBR, na.rm = TRUE),
    avg_speed_limit = mean(SPEED_LIMIT_NBR, na.rm = TRUE),
    avg_PCR = mean(PCR_NBR, na.rm = TRUE),
    avg_crack_deduct = mean(TOT_CRACK_DEDUCT_NBR, na.rm = TRUE),
    avg_rut_deduct = mean(TOT_RUT_DEDUCT_NBR, na.rm = TRUE),
    avg_lanes = mean(LANES_NBR, na.rm = TRUE),
    avg_baseline_crash_rate = mean(crashes_per_MVMT, na.rm = TRUE),
    n = n()
  ) %>%
  as.data.frame()

# Bin segments by their baseline crash rate and see improvement rates
rf_df %>%
  mutate(
    baseline_bin = cut(crashes_per_MVMT,
                       breaks = c(0, 10, 25, 50, 100, Inf),
                       labels = c("Very Low (0-10)", "Low (10-25)", 
                                  "Medium (25-50)", "High (50-100)", 
                                  "Very High (100+)"))
  ) %>%
  group_by(baseline_bin) %>%
  summarise(
    n_segments = n(),
    pct_improved = mean(improved == "Improved") * 100,
    avg_change = mean(as.numeric(as.character(
      factor(improved, levels = c("Improved","Worsened"), labels = c(-1,1))
    )), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as.data.frame()

# Summary plot of all DiD effects including the "all projects" result
all_effects <- did_by_category %>%
  select(work_category, did_effect) %>%
  bind_rows(
    data.frame(
      work_category = "All Safety Projects (Overall)",
      did_effect = did_safety_effect_value
    )
  ) %>%
  mutate(
    significant = work_category == "Intersection Improvement (Safety)",
    work_category = str_wrap(work_category, width = 30)
  )

ggplot(all_effects, aes(x = reorder(work_category, did_effect),
                        y = did_effect,
                        fill = did_effect < 0)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_text(aes(label = round(did_effect, 2)),
            hjust = ifelse(all_effects$did_effect < 0, 1.2, -0.2),
            size = 5.5) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Crash reduction", "FALSE" = "Crash increase")
  ) +
  labs(
    title = "Which Safety Projects Actually Reduce Crashes?",
    subtitle = "Difference-in-Differences effect vs matched control segments",
    x = NULL,
    y = "DiD Effect (crashes per MVMT)",
    fill = NULL,
    caption = "Negative values indicate crash reduction relative to control roads"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")





### High Risk Model


# Build a high risk indicator and profile what those segments look like
high_risk_profile <- analysis_df %>%
  st_drop_geometry() %>%
  mutate(
    high_risk = crashes_per_MVMT >= quantile(crashes_per_MVMT, 0.90, na.rm = TRUE),
    received_safety_investment = ROADWAY_INVENTORY_ID %in% 
      treated_safety_ids$ROADWAY_INVENTORY_ID
  ) %>%
  group_by(high_risk) %>%
  summarise(
    n_segments = n(),
    avg_ADT = mean(ADT_TOTAL_NBR, na.rm = TRUE),
    avg_speed_limit = mean(SPEED_LIMIT_NBR, na.rm = TRUE),
    avg_PCR = mean(PCR_NBR, na.rm = TRUE),
    avg_length = mean(SEGMENT_LENGTH_NBR, na.rm = TRUE),
    avg_lanes = mean(LANES_NBR, na.rm = TRUE),
    avg_IRI = mean(AVG_IRI, na.rm = TRUE),
    avg_crack_deduct = mean(TOT_CRACK_DEDUCT_NBR, na.rm = TRUE),
    avg_rut_deduct = mean(TOT_RUT_DEDUCT_NBR, na.rm = TRUE),
    pct_received_investment = mean(received_safety_investment) * 100,
    .groups = "drop"
  )

as.data.frame(high_risk_profile)

# Also look at functional class breakdown of high risk segments
analysis_df %>%
  st_drop_geometry() %>%
  mutate(high_risk = crashes_per_MVMT >= quantile(crashes_per_MVMT, 0.90, na.rm = TRUE)) %>%
  group_by(FUNCTION_CLASS_CD, high_risk) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = high_risk, values_from = n, names_prefix = "high_risk_") %>%
  mutate(pct_high_risk = high_risk_TRUE / (high_risk_TRUE + high_risk_FALSE) * 100) %>%
  arrange(desc(pct_high_risk)) %>%
  as.data.frame()






# Build severity profile by functional class and risk level
severity_profile <- analysis_df %>%
  st_drop_geometry() %>%
  mutate(
    high_risk = crashes_per_MVMT >= quantile(crashes_per_MVMT, 0.90, na.rm = TRUE)
  ) %>%
  group_by(FUNCTION_CLASS_CD, high_risk) %>%
  summarise(
    n_segments = n(),
    total_crashes = sum(total_crashes, na.rm = TRUE),
    total_fatal = sum(fatal_crashes, na.rm = TRUE),
    total_injury = sum(injury_crashes, na.rm = TRUE),
    fatal_rate = sum(fatal_crashes) / sum(total_crashes) * 100,
    injury_rate = sum(injury_crashes) / sum(total_crashes) * 100,
    crashes_per_segment = mean(total_crashes, na.rm = TRUE),
    avg_ADT = mean(ADT_TOTAL_NBR, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(FUNCTION_CLASS_CD, high_risk) %>%
  as.data.frame()

print(severity_profile)

# Also get a clean high risk vs normal summary by severity
analysis_df %>%
  st_drop_geometry() %>%
  mutate(
    high_risk = crashes_per_MVMT >= quantile(crashes_per_MVMT, 0.90, na.rm = TRUE)
  ) %>%
  group_by(high_risk) %>%
  summarise(
    total_crashes = sum(total_crashes, na.rm = TRUE),
    total_fatal = sum(fatal_crashes, na.rm = TRUE),
    total_injury = sum(injury_crashes, na.rm = TRUE),
    pct_of_all_fatals = NA,  # will compute below
    fatal_rate = sum(fatal_crashes) / sum(total_crashes) * 100,
    injury_rate = sum(injury_crashes) / sum(total_crashes) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    pct_of_all_fatals = total_fatal / sum(total_fatal) * 100,
    pct_of_all_injuries = total_injury / sum(total_injury) * 100,
    pct_of_all_crashes = total_crashes / sum(total_crashes) * 100
  ) %>%
  as.data.frame()
```

```{r}
#############################
### What is more dangerous? 
#############################

fc_summary <- analysis_df %>%
  st_drop_geometry() %>%
  group_by(FUNCTION_CLASS_CD) %>%
  summarise(
    n_segments = n(),
    avg_ADT = mean(ADT_TOTAL_NBR, na.rm = TRUE),
    total_crashes = sum(total_crashes, na.rm = TRUE),
    total_fatal = sum(fatal_crashes, na.rm = TRUE),
    total_injury = sum(injury_crashes, na.rm = TRUE),
    avg_crashes_per_MVMT = round(mean(crashes_per_MVMT, na.rm = TRUE), 1),
    fatal_rate_pct = round(sum(fatal_crashes) / sum(total_crashes) * 100, 2),
    .groups = "drop"
  ) %>%
  mutate(
    pct_of_all_crashes = round(total_crashes / sum(total_crashes) * 100, 1),
    pct_of_all_fatals = round(total_fatal / sum(total_fatal) * 100, 1),
    avg_crashes_per_segment = round(total_crashes / n_segments, 1)
  ) %>%
  arrange(FUNCTION_CLASS_CD)

as.data.frame(fc_summary)

analysis_df %>%
  st_drop_geometry() %>%
  mutate(
    received_safety = ROADWAY_INVENTORY_ID %in% treated_safety_ids$ROADWAY_INVENTORY_ID
  ) %>%
  filter(received_safety == TRUE) %>%
  group_by(FUNCTION_CLASS_CD) %>%
  summarise(
    n_treated_segments = n(),
    avg_ADT = mean(ADT_TOTAL_NBR, na.rm = TRUE),
    avg_crashes_per_MVMT = round(mean(crashes_per_MVMT, na.rm = TRUE), 1),
    total_fatal = sum(fatal_crashes, na.rm = TRUE),
    fatal_rate_pct = round(sum(fatal_crashes) / sum(total_crashes) * 100, 2),
    .groups = "drop"
  ) %>%
  mutate(
    pct_of_safety_work = round(n_treated_segments / sum(n_treated_segments) * 100, 1)
  ) %>%
  arrange(desc(n_treated_segments)) %>%
  as.data.frame()

# Build the investment vs need comparison directly
investment_vs_need <- fc_summary %>%
  left_join(
    analysis_df %>%
      st_drop_geometry() %>%
      mutate(received_safety = ROADWAY_INVENTORY_ID %in% 
               treated_safety_ids$ROADWAY_INVENTORY_ID) %>%
      group_by(FUNCTION_CLASS_CD) %>%
      summarise(
        n_treated = sum(received_safety),
        pct_treated = round(mean(received_safety) * 100, 1),
        .groups = "drop"
      ),
    by = "FUNCTION_CLASS_CD"
  ) %>%
  mutate(
    fc_label = case_when(
      FUNCTION_CLASS_CD == 1 ~ "1 - Interstate",
      FUNCTION_CLASS_CD == 2 ~ "2 - Freeway",
      FUNCTION_CLASS_CD == 3 ~ "3 - Major Arterial",
      FUNCTION_CLASS_CD == 4 ~ "4 - Minor Arterial",
      FUNCTION_CLASS_CD == 5 ~ "5 - Collector",
      FUNCTION_CLASS_CD == 6 ~ "6 - Minor Collector",
      FUNCTION_CLASS_CD == 7 ~ "7 - Local"
    )
  ) %>%
  select(fc_label, n_segments, pct_treated, avg_crashes_per_MVMT, 
         pct_of_all_fatals, fatal_rate_pct, avg_ADT)

as.data.frame(investment_vs_need)
```

```{r}
############################################
### Cincinnati / Hamilton County Analysis
############################################

# Check what county identifier looks like in your data
unique(analysis_df$COUNTY_CD.x) %>% head(20)

# Hamilton County code in Ohio is 31
# Correct filter using abbreviation
cincinnati_df <- analysis_df %>%
  st_drop_geometry() %>%
  filter(COUNTY_CD.x == "HAM")

nrow(cincinnati_df)

# Verify it worked
head(cincinnati_df$COUNTY_CD.x)

# Also filter segment_year_data to Cincinnati
cincinnati_year <- segment_year_data %>%
  filter(ROADWAY_INVENTORY_ID %in% cincinnati_df$ROADWAY_INVENTORY_ID)

# Quick profile of Cincinnati roads
cincinnati_df %>%
  group_by(FUNCTION_CLASS_CD) %>%
  summarise(
    n_segments = n(),
    avg_ADT = mean(ADT_TOTAL_NBR, na.rm = TRUE),
    avg_crash_rate = mean(crashes_per_MVMT, na.rm = TRUE),
    total_fatal = sum(fatal_crashes, na.rm = TRUE),
    pct_treated = mean(ROADWAY_INVENTORY_ID %in% 
                         treated_safety_ids$ROADWAY_INVENTORY_ID) * 100,
    .groups = "drop"
  ) %>%
  arrange(FUNCTION_CLASS_CD) %>%
  as.data.frame()



COL_ORANGE  <- "#f0883e"
COL_BLUE    <- "#58a6ff"
COL_GREEN   <- "#3fb950"
COL_RED     <- "#f85149"
COL_YELLOW  <- "#d29922"
COL_MUTED   <- "#8b949e"
COL_SURFACE <- "#161b22"

theme_odot <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.background    = element_rect(fill = "#0d1117", color = NA),
      panel.background   = element_rect(fill = "#161b22", color = NA),
      panel.grid.major   = element_line(color = "#30363d", linewidth = 0.4),
      panel.grid.minor   = element_blank(),
      panel.border       = element_rect(color = "#30363d", fill = NA),
      text               = element_text(color = "#e6edf3"),
      axis.text          = element_text(color = "#8b949e", size = 10),
      axis.title         = element_text(color = "#e6edf3", size = 11),
      plot.title         = element_text(color = "#e6edf3", size = 14,
                                        face = "bold",
                                        margin = ggplot2::margin(b = 4)),
      plot.subtitle      = element_text(color = "#8b949e", size = 10,
                                        margin = ggplot2::margin(b = 16)),
      plot.caption       = element_text(color = "#8b949e", size = 9,
                                        margin = ggplot2::margin(t = 12)),
      legend.background  = element_rect(fill = "#161b22", color = NA),
      legend.text        = element_text(color = "#8b949e", size = 10),
      legend.title       = element_text(color = "#e6edf3", size = 10),
      strip.text         = element_text(color = "#e6edf3", size = 10),
      plot.margin        = ggplot2::margin(16, 16, 16, 16)
    )
}

cincinnati_fc_plot <- data.frame(
  fc_label      = factor(c("1-Interstate", "2-Freeway", "3-Major Art.",
                           "4-Minor Art.", "5-Collector", "7-Local"),
                         levels = c("1-Interstate", "2-Freeway", "3-Major Art.",
                                    "4-Minor Art.", "5-Collector", "7-Local")),
  avg_crash_rate = c(18.66, 11.88, 33.84, 37.29, 40.23, 41.23),
  pct_treated    = c(4.78, 6.35, 14.16, 6.52, 2.18, 0.42),
  total_fatal    = c(39, 4, 75, 69, 47, 15),
  n_segments     = c(691, 189, 883, 966, 1146, 710)
) %>%
  mutate(
    bar_color = case_when(
      fc_label %in% c("1-Interstate", "2-Freeway") ~ COL_BLUE,
      fc_label == "3-Major Art."                   ~ COL_YELLOW,
      fc_label %in% c("4-Minor Art.", "5-Collector") ~ COL_RED,
      fc_label == "7-Local"                         ~ COL_ORANGE
    ),
    pct_of_fatals = round(total_fatal / sum(total_fatal) * 100, 1)
  )

# ── Plot A: Investment Rate vs Crash Rate (bubble) ───────────────────────────
p_bubble <- ggplot(cincinnati_fc_plot,
                   aes(x = pct_treated, y = avg_crash_rate,
                       size = pct_of_fatals, fill = bar_color)) +
  geom_point(shape = 21, alpha = 0.85,
             color = "#0d1117", stroke = 1.5) +
  geom_text(aes(label = fc_label),
            vjust = -1.5, size = 5.5, color = "#e6edf3") +
  annotate("rect", xmin = -0.3, xmax = 4,
           ymin = 35, ymax = 44,
           fill = COL_RED, alpha = 0.06,
           color = COL_RED, linetype = "dashed", linewidth = 0.4) +
  annotate("text", x = 1.8, y = 43.5,
           label = "High risk, low investment",
           color = COL_RED, size = 5.0, fontface = "italic") +
  scale_fill_identity() +
  scale_size_continuous(range = c(5, 20),
                        name = "% of Local Fatalities") +
  scale_x_continuous(limits = c(-0.5, 17),
                     labels = scales::percent_format(scale = 1)) +
  labs(
    title    = "Cincinnati: Investment Gap by Road Class",
    subtitle = "Bubble size = share of Hamilton County fatalities",
    x        = "% of Segments Receiving Safety Investment",
    y        = "Avg Crash Rate (per MVMT)"
  ) +
  theme_odot() +
  theme(legend.position = "right")

p_bubble

# ── Plot B: Side by side investment vs crash rate bars ───────────────────────
p_bars <- cincinnati_fc_plot %>%
  select(fc_label, avg_crash_rate, pct_treated) %>%
  pivot_longer(-fc_label,
               names_to  = "metric",
               values_to = "value") %>%
  mutate(metric = case_when(
    metric == "avg_crash_rate" ~ "Avg Crash Rate per MVMT",
    metric == "pct_treated"    ~ "% Segments Receiving Safety Work"
  )) %>%
  ggplot(aes(x = fc_label, y = value, fill = metric)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.85) +
  scale_fill_manual(values = c(
    "Avg Crash Rate per MVMT"          = COL_ORANGE,
    "% Segments Receiving Safety Work" = COL_BLUE
  )) +
  labs(
    title    = "Risk vs Investment — Hamilton County",
    subtitle = "Collectors and local roads are the most underserved",
    x = NULL, y = "Value", fill = NULL
  ) +
  theme_odot() +
  theme(
    legend.position = "bottom",
    axis.text.x     = element_text(angle = 30, hjust = 1)
  )

p_bars

# ── Plot C: Fatal share bar ─────────────────────────────────────────────────
p_fatal <- ggplot(cincinnati_fc_plot,
                  aes(x = reorder(fc_label, -pct_of_fatals),
                      y = pct_of_fatals,
                      fill = bar_color)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(aes(label = paste0(pct_of_fatals, "%")),
            vjust = -0.5, size = 5.5, color = "#e6edf3") +
  scale_fill_identity() +
  scale_y_continuous(limits = c(0, 40),
                     labels = scales::percent_format(scale = 1)) +
  labs(
    title    = "Hamilton County Fatalities by Road Class",
    subtitle = "% of all Hamilton County traffic fatalities 2020–2024",
    x        = NULL,
    y        = "% of Local Fatalities"
  ) +
  theme_odot()

p_fatal

# ── Combined panel ────────────────────────────────────────────────────────────
library(patchwork)

cincinnati_panel <- (p_bubble | p_fatal) / p_bars +
  plot_annotation(
    title    = "Cincinnati (Hamilton County) — Safety Investment Analysis",
    subtitle = "4,585 road segments | 249 total fatalities 2020–2024",
    caption  = "Source: ODOT Crash Data 2020–2024 · Road Inventory",
    theme    = theme(
      plot.background = element_rect(fill = "#0d1117", color = NA),
      plot.title      = element_text(color = "#e6edf3", size = 16, face = "bold"),
      plot.subtitle   = element_text(color = "#8b949e", size = 11),
      plot.caption    = element_text(color = "#8b949e", size = 9)
    )
  )

cincinnati_panel
```

```{r}
#####################
### Testing DID
#####################

# Test parallel trends using pre-period years only
# We need yearly crash rates for 2020 and 2021 (both "before" years)

parallel_trends_test <- segment_year_data %>%
  filter(CRASH_YR %in% c(2020, 2021)) %>%
  group_by(treated_safety, CRASH_YR) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    .groups = "drop"
  )

print(parallel_trends_test)

# Visualize
ggplot(parallel_trends_test,
       aes(x = CRASH_YR, y = avg_crash_rate,
           color = treated_safety, group = treated_safety)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3) +
  scale_color_manual(values = c("FALSE" = COL_MUTED, "TRUE" = COL_BLUE),
                     labels = c("FALSE" = "Control", "TRUE" = "Treated")) +
  scale_x_continuous(breaks = c(2020, 2021)) +
  labs(
    title    = "Parallel Trends Test: Pre-Treatment Period",
    subtitle = "Lines should be parallel for DiD assumption to hold",
    x        = "Year",
    y        = "Avg Crash Rate per MVMT",
    color    = NULL
  ) +
  theme_odot()




# Check how many segments appear in both before and after
period_counts <- segment_year_data %>%
  group_by(ROADWAY_INVENTORY_ID) %>%
  summarise(
    has_before = any(period == "before"),
    has_after  = any(period == "after"),
    .groups = "drop"
  )

table(period_counts$has_before, period_counts$has_after,
      dnn = c("Has Before", "Has After"))


# Compare raw vs log-transformed crash rates
segment_year_data %>%
  filter(crash_rate_MVMT > 0, crash_rate_MVMT < 500) %>%
  mutate(log_rate = log(crash_rate_MVMT)) %>%
  pivot_longer(cols = c(crash_rate_MVMT, log_rate),
               names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 80, fill = COL_BLUE, alpha = 0.7) +
  facet_wrap(~metric, scales = "free") +
  labs(
    title    = "Crash Rate Distribution: Raw vs Log Transformed",
    subtitle = "Log transformation may better satisfy model assumptions",
    x = NULL, y = "Count"
  ) +
  theme_odot()

# If log looks more normal, rerun DiD with log outcome
balanced_segments <- balanced_segments %>%
  mutate(log_crash_rate = log(crash_rate_MVMT + 0.01))

did_model_log <- felm(
  log_crash_rate ~ treated_safety * I(period == "after") |
    ROADWAY_INVENTORY_ID + CRASH_YR |
    0 |
    ROADWAY_INVENTORY_ID,
  data = balanced_segments %>% filter(is.finite(log_crash_rate))
)

summary(did_model_log)




### RESULTS

## Parralell Line test shows no extreme converging or diverging trend so that fine

## 
```

```{r}
############################################
### ODOT Safety Investment Analysis
### Progress Report Visualizations
### Xavier University Senior Capstone 2026
############################################

library(tidyverse)
library(ggplot2)
library(patchwork)  # install.packages("patchwork") if needed

# ── Shared theme ──────────────────────────────────────────────────────────────


theme_odot <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.background    = element_rect(fill = "#0d1117", color = NA),
      panel.background   = element_rect(fill = "#161b22", color = NA),
      panel.grid.major   = element_line(color = "#30363d", linewidth = 0.4),
      panel.grid.minor   = element_blank(),
      panel.border       = element_rect(color = "#30363d", fill = NA),
      text               = element_text(color = "#e6edf3"),
      axis.text          = element_text(color = "#8b949e", size = 10),
      axis.title         = element_text(color = "#e6edf3", size = 11),
      plot.title         = element_text(color = "#e6edf3", size = 14,
                                        face = "bold",
                                        margin = ggplot2::margin(b = 4)),
      plot.subtitle      = element_text(color = "#8b949e", size = 10,
                                        margin = ggplot2::margin(b = 16)),
      plot.caption       = element_text(color = "#8b949e", size = 9,
                                        margin = ggplot2::margin(t = 12)),
      legend.background  = element_rect(fill = "#161b22", color = NA),
      legend.text        = element_text(color = "#8b949e", size = 10),
      legend.title       = element_text(color = "#e6edf3", size = 10),
      strip.text         = element_text(color = "#e6edf3", size = 10),
      plot.margin        = ggplot2::margin(16, 16, 16, 16)
    )
}

# Color palette
COL_ORANGE  <- "#f0883e"
COL_BLUE    <- "#58a6ff"
COL_GREEN   <- "#3fb950"
COL_RED     <- "#f85149"
COL_YELLOW  <- "#d29922"
COL_MUTED   <- "#8b949e"
COL_SURFACE <- "#161b22"


############################################
### FIGURE 1: Difference-in-Differences
### Before vs After — Matched Control
############################################

did_plot_data <- data.frame(
  period  = factor(c("Before", "After", "Before", "After"),
                   levels = c("Before", "After")),
  group   = c("Control (Matched ADT)",  "Control (Matched ADT)",
              "Treated (Safety-Funded)", "Treated (Safety-Funded)"),
  rate    = c(10.6, 10.3, 7.65, 7.42)
)

fig1 <- ggplot(did_plot_data,
               aes(x = period, y = rate, color = group, group = group)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 4) +
  geom_label(aes(label = round(rate, 2)),
             nudge_x = 0.07, show.legend = FALSE,
             fill = "#161b22", size = 5.5,
             linewidth = 0)  + # <-- replaces label.size = 0
  scale_color_manual(values = c("Control (Matched ADT)"   = COL_MUTED,
                                "Treated (Safety-Funded)" = COL_BLUE)) +
  labs(
    title    = "Difference-in-Differences: Safety Project Impact",
    subtitle = "Segments matched on ADT volume | DiD effect = +0.07 (p = 0.908, not significant)",
    x        = NULL,
    y        = "Average Crash Rate (per MVMT)",
    color    = NULL,
    caption  = "Source: ODOT Crash Data 2020–2024"
  ) +
  theme_odot() +
  theme(legend.position = "bottom")

fig1


############################################
### FIGURE 2: DiD Effect by Project Type
############################################

category_data <- data.frame(
  category   = c("Intersection Imp. (Safety)",
                 "Pedestrian Facilities",
                 "Roadway Minor Rehab",
                 "Roadside/Median Imp.",
                 "Pavement Treatments (Safety)",
                 "Traffic Control (Safety)"),
  did_effect = c(-6.54, 0.14, 0.34, 1.15, 1.65, 2.49),
  n_segments = c(697, 277, 1266, 664, 263, 2739)
)

fig2 <- ggplot(category_data,
               aes(x = reorder(category, -did_effect),
                   y = did_effect,
                   fill = did_effect < 0)) +
  geom_col(width = 0.65) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = COL_MUTED, linewidth = 0.7) +
  geom_text(aes(label = round(did_effect, 2),
                hjust = ifelse(did_effect < 0, 1.3, -0.3)),
            color = "#e6edf3", size = 5.5) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = COL_GREEN, "FALSE" = COL_RED),
                    labels = c("TRUE" = "Crash reduction",
                               "FALSE" = "Crash increase")) +
  labs(
    title    = "Which Safety Projects Actually Reduce Crashes?",
    subtitle = "DiD effect vs matched control segments | Negative = crash reduction",
    x        = NULL,
    y        = "DiD Effect (crashes per MVMT vs control)",
    fill     = NULL,
    caption  = "Only categories with 200+ treated segments shown"
  ) +
  theme_odot() +
  theme(legend.position = "bottom")

fig2


############################################
### FIGURE 3: Event Study — All Projects
############################################

event_data <- data.frame(
  event_time     = -2:2,
  avg_crash_rate = c(8.10, 7.64, 6.74, 6.80, 7.42),
  se             = c(0.755, 0.689, 0.428, 0.409, 0.532)
)

fig3 <- ggplot(event_data, aes(x = event_time, y = avg_crash_rate)) +
  geom_ribbon(aes(ymin = avg_crash_rate - 1.96 * se,
                  ymax = avg_crash_rate + 1.96 * se),
              fill = COL_BLUE, alpha = 0.15) +
  geom_line(linewidth = 1.3, color = COL_BLUE) +
  geom_point(size = 4, color = COL_BLUE) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = COL_RED, linewidth = 0.8) +
  annotate("text", x = 0.15, y = 8.3,
           label = "Construction\nComplete",
           color = COL_RED, hjust = 0, size = 5.5) +
  scale_x_continuous(breaks = -2:2) +
  labs(
    title    = "Event Study: Crash Rates Around Safety Project Completion",
    subtitle = "Treated segments only | t=0 is completion year | 95% CI shown",
    x        = "Years Relative to Project Completion",
    y        = "Average Crash Rate (per MVMT)",
    caption  = "n = 2,046–2,662 segments per time point"
  ) +
  theme_odot()

fig3


############################################
### FIGURE 4: 2021 Cohort — Lag Test
############################################

cohort_data <- data.frame(
  event_time     = -1:3,
  avg_crash_rate = c(6.5, 6.9, 7.2, 8.3, 9.0)
)

fig4 <- ggplot(cohort_data, aes(x = event_time, y = avg_crash_rate)) +
  geom_area(fill = COL_RED, alpha = 0.1) +
  geom_line(linewidth = 1.3, color = COL_RED) +
  geom_point(size = 4, color = COL_RED) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = COL_MUTED, linewidth = 0.8) +
  annotate("text", x = 0.15, y = 9.2,
           label = "Construction\nComplete",
           color = COL_MUTED, hjust = 0, size = 5.5) +
  scale_x_continuous(breaks = -1:3) +
  labs(
    title    = "2021 Cohort: No Lag Effect Found",
    subtitle = "Crash rates increase steadily post-completion — no delayed safety benefit",
    x        = "Years Relative to 2021 Project Completion",
    y        = "Average Crash Rate (per MVMT)",
    caption  = "Only segments completing construction in 2021 | Extended to t=+3"
  ) +
  theme_odot()

fig4


############################################
### FIGURE 5: Fatalities by Functional Class
############################################

fc_data <- data.frame(
  fc_label      = factor(c("1-Interstate","2-Freeway","3-Major Art.",
                           "4-Minor Art.","5-Collector","6-Minor Coll.","7-Local"),
                         levels = c("1-Interstate","2-Freeway","3-Major Art.",
                                    "4-Minor Art.","5-Collector",
                                    "6-Minor Coll.","7-Local")),
  pct_fatals    = c(8.1, 2.3, 19.3, 24.0, 28.5, 4.9, 12.8),
  crash_rate    = c(11.3, 8.0, 23.7, 25.6, 25.0, 22.6, 36.5),
  fatal_rate    = c(0.47, 0.48, 0.40, 0.43, 0.58, 0.82, 0.58),
  pct_treated   = c(9.8, 12.0, 14.6, 6.1, 3.6, 4.0, 1.2),
  avg_adt       = c(29301, 14841, 13202, 8359, 3722, 1111, 1034)
)

# Highlight FC 4+5 as the key finding
fc_data <- fc_data %>%
  mutate(
    highlight = fc_label %in% c("4-Minor Art.", "5-Collector"),
    bar_color = case_when(
      fc_label %in% c("1-Interstate", "2-Freeway") ~ COL_BLUE,
      fc_label %in% c("4-Minor Art.", "5-Collector") ~ COL_RED,
      fc_label == "7-Local" ~ COL_ORANGE,
      TRUE ~ COL_YELLOW
    )
  )

fig5 <- ggplot(fc_data, aes(x = fc_label)) +
  geom_col(aes(y = pct_fatals, fill = bar_color),
           width = 0.6, alpha = 0.85) +
  geom_line(aes(y = crash_rate, group = 1),
            color = "white", linewidth = 1.2) +
  geom_point(aes(y = crash_rate),
             color = "white", size = 5.5) +
  scale_fill_identity() +
  scale_y_continuous(
    name     = "% of All Fatalities (bars)",
    sec.axis = sec_axis(~., name = "Avg Crash Rate per MVMT (line)")
  ) +
  annotate("rect", xmin = 3.55, xmax = 5.45,
           ymin = -1, ymax = 31,
           fill = COL_RED, alpha = 0.06) +
  annotate("text", x = 4.5, y = 33,
           label = "52.5% of all\nfatalities",
           color = COL_RED, size = 5.5, fontface = "bold") +
  labs(
    title    = "Where Is the Harm? Fatalities & Crash Rates by Road Class",
    subtitle = "FC 4 + FC 5 roads account for more than half of all Ohio traffic fatalities",
    x        = NULL,
    caption  = "Source: ODOT Crash Data 2020–2024 · 4,466 total fatalities"
  ) +
  theme_odot() +
  theme(axis.title.y.right = element_text(color = "white"))

fig5


#Minor Arterial roads are a bridge between high-speed highways and local 
#neighborhood streets. They are designed to carry moderate-length trips through 
#a region at slower speeds

#Collector roads are low-to-moderate capacity roads that "collect" traffic from 
#local residential streets and funnel it onto major arterials or highways. 

############################################
### FIGURE 6: The Investment Gap
### % Treated vs Crash Rate (bubble = fatality share)
############################################

fig6 <- ggplot(fc_data,
               aes(x = pct_treated, y = crash_rate,
                   size = pct_fatals, fill = bar_color)) +
  geom_point(shape = 21, alpha = 0.8, color = "#0d1117", stroke = 1.5) +
  geom_text(aes(label = fc_label),
            vjust = -1.4, size = 5.5, color = "#e6edf3") +
  # Danger zone annotation
  annotate("rect", xmin = -0.5, xmax = 5,
           ymin = 28, ymax = 40,
           fill = COL_RED, alpha = 0.05,
           linetype = "dashed", color = COL_RED, linewidth = 0.5) +
  annotate("text", x = 2.5, y = 39,
           label = "High risk, low investment",
           color = COL_RED, size = 3, fontface = "italic") +
  scale_fill_identity() +
  scale_size_continuous(range = c(4, 18), name = "% of All Fatalities") +
  scale_x_continuous(limits = c(-0.5, 17),
                     labels = scales::percent_format(scale = 1)) +
  labs(
    title    = "The Investment Gap: Where Safety Funding Goes vs Where It's Needed",
    subtitle = "Bubble size = share of all fatalities | Local roads sit in the danger zone",
    x        = "% of Segments Receiving Safety Investment",
    y        = "Average Crash Rate (per MVMT)",
    caption  = "Ideal: high investment AND low crash rate (upper left = over-invested, lower right = under-invested)"
  ) +
  theme_odot() +
  theme(legend.position = "right")

fig6


############################################
### FIGURE 7: Investment Gap Bar Chart
### Side-by-side: % treated vs crash rate
############################################

gap_long <- fc_data %>%
  select(fc_label, pct_treated, crash_rate) %>%
  pivot_longer(cols = c(pct_treated, crash_rate),
               names_to = "metric", values_to = "value") %>%
  mutate(
    metric = case_when(
      metric == "pct_treated" ~ "% Segments Receiving Safety Work",
      metric == "crash_rate"  ~ "Avg Crash Rate per MVMT"
    )
  )

fig7 <- ggplot(gap_long,
               aes(x = fc_label, y = value,
                   fill = metric)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.85) +
  scale_fill_manual(values = c(
    "% Segments Receiving Safety Work" = COL_BLUE,
    "Avg Crash Rate per MVMT"          = COL_ORANGE
  )) +
  labs(
    title    = "Safety Investment Rate vs Road Risk by Functional Class",
    subtitle = "Investment (blue) vs crash rate (orange) — gap is largest for local roads",
    x        = NULL,
    y        = "Value",
    fill     = NULL,
    caption  = "Local roads (FC 7): 1.2% investment rate despite highest crash rate (36.5/MVMT)"
  ) +
  theme_odot() +
  theme(
    legend.position  = "bottom",
    axis.text.x      = element_text(angle = 30, hjust = 1)
  )

fig7


############################################
### FIGURE 8: High Risk Road Profile
### Improved vs Worsened segments
############################################

profile_data <- data.frame(
  group    = c("Improved", "Worsened"),
  avg_adt  = c(11666, 11086),
  avg_pcr  = c(81.6, 82.1),
  avg_spd  = c(40.6, 41.6),
  baseline = c(39.1, 31.4),
  n        = c(15667, 23473)
)

profile_long <- profile_data %>%
  select(group, avg_adt, avg_pcr, avg_spd, baseline) %>%
  pivot_longer(-group, names_to = "metric", values_to = "value") %>%
  mutate(metric = case_when(
    metric == "avg_adt"  ~ "Avg ADT (Traffic Volume)",
    metric == "avg_pcr"  ~ "Avg PCR (Pavement Score)",
    metric == "avg_spd"  ~ "Avg Speed Limit (mph)",
    metric == "baseline" ~ "Baseline Crash Rate (per MVMT)"
  ))

fig8 <- ggplot(profile_long,
               aes(x = group, y = value, fill = group)) +
  geom_col(width = 0.55, alpha = 0.85) +
  facet_wrap(~metric, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = c("Improved" = COL_BLUE,
                               "Worsened" = COL_RED)) +
  labs(
    title    = "Road Characteristics: Improved vs Worsened Segments",
    subtitle = "Baseline crash rate is the key differentiator — all other characteristics nearly identical",
    x        = NULL,
    y        = NULL,
    fill     = NULL,
    caption  = "Improved n=15,667 | Worsened n=23,473 | Improvement defined as crash rate decline before→after"
  ) +
  theme_odot() +
  theme(
    legend.position = "none",
    strip.text      = element_text(size = 9, color = "#8b949e")
  )

fig8
```

```{r}
############################################
### Case Study: Find Candidate Projects
############################################

case_study_candidates <- analysis_df %>%
  st_drop_geometry() %>%
  inner_join(
    projects_safety %>%
      select(NLF_ID, CTL_BEGIN, CTL_END, completion_year,
             PRIMARY_WORK_CATEGORY, EST_TOTAL_CONSTR_COST,
             COUNTY_NME, DISTRICT_NBR, PROJECT_NME,
             END_CONSTR_MILESTONE_DT),
    by = "NLF_ID",
    relationship = "many-to-many"
  ) %>%
  filter(
    CTL_BEGIN_NBR <= CTL_END.y,  # project end >= segment start
    CTL_END_NBR >= CTL_BEGIN.y   # project start <= segment end
  ) %>%
  filter(
    PRIMARY_WORK_CATEGORY %in% c(
      "Intersection Improvement (Safety)",
      "Traffic Control (Safety)",
      "Roadway Minor Rehab"
    )
  ) %>%
  left_join(
    segment_year_data %>%
      group_by(ROADWAY_INVENTORY_ID) %>%
      summarise(
        before_rate = mean(crash_rate_MVMT[period == "before"], na.rm = TRUE),
        after_rate  = mean(crash_rate_MVMT[period == "after"],  na.rm = TRUE),
        change      = after_rate - before_rate,
        .groups     = "drop"
      ),
    by = "ROADWAY_INVENTORY_ID"
  ) %>%
  filter(!is.na(change)) %>%
  select(ROADWAY_INVENTORY_ID, NLF_ID, PROJECT_NME, COUNTY_NME,
         COUNTY_CD.x, PRIMARY_WORK_CATEGORY, completion_year,
         EST_TOTAL_CONSTR_COST, FUNCTION_CLASS_CD, ADT_TOTAL_NBR,
         SPEED_LIMIT_NBR, before_rate, after_rate, change)

# Check Hamilton County availability
case_study_candidates %>%
  filter(COUNTY_CD.x == "HAM") %>%
  count(PRIMARY_WORK_CATEGORY, sort = TRUE) %>%
  as.data.frame()
```

```{r}
# Best intersection improvement — biggest crash reduction
best_intersection <- case_study_candidates %>%
  filter(COUNTY_CD.x == "HAM",
         PRIMARY_WORK_CATEGORY == "Intersection Improvement (Safety)") %>%
  slice_min(change, n = 3) %>%
  select(ROADWAY_INVENTORY_ID, PROJECT_NME, COUNTY_NME,
         completion_year, before_rate, after_rate, change,
         EST_TOTAL_CONSTR_COST, FUNCTION_CLASS_CD, ADT_TOTAL_NBR)

# Best traffic control — biggest crash increase (most interesting negative case)
best_traffic <- case_study_candidates %>%
  filter(COUNTY_CD.x == "HAM",
         PRIMARY_WORK_CATEGORY == "Traffic Control (Safety)") %>%
  slice_max(change, n = 3) %>%
  select(ROADWAY_INVENTORY_ID, PROJECT_NME, COUNTY_NME,
         completion_year, before_rate, after_rate, change,
         EST_TOTAL_CONSTR_COST, FUNCTION_CLASS_CD, ADT_TOTAL_NBR)

# Best rehab — closest to zero change (true null result)
best_rehab <- case_study_candidates %>%
  filter(COUNTY_CD.x == "HAM",
         PRIMARY_WORK_CATEGORY == "Roadway Minor Rehab") %>%
  mutate(abs_change = abs(change)) %>%
  slice_min(abs_change, n = 3) %>%
  select(ROADWAY_INVENTORY_ID, PROJECT_NME, COUNTY_NME,
         completion_year, before_rate, after_rate, change,
         EST_TOTAL_CONSTR_COST, FUNCTION_CLASS_CD, ADT_TOTAL_NBR)

cat("=== INTERSECTION IMPROVEMENT CANDIDATES ===\n")
as.data.frame(best_intersection)

cat("\n=== TRAFFIC CONTROL CANDIDATES ===\n")
as.data.frame(best_traffic)

cat("\n=== ROADWAY REHAB CANDIDATES ===\n")
as.data.frame(best_rehab)
```

```{r}
############################################
### Case Study Profiles
############################################

# Define our three chosen segments
case_ids <- c(
  "2024_000000000018030",  # Intersection - Plainfield Roundabouts
  "2024_000000000271525",  # Traffic Control - Eastern Corridor
  "2024_000000000270710"   # Rehab - SR 32/SR 125
)

case_labels <- c(
  "2024_000000000018030" = "Plainfield Rd Roundabouts\n(Intersection Improvement)",
  "2024_000000000271525" = "Eastern Corridor VAR TSG\n(Traffic Control)",
  "2024_000000000270710" = "SR 32/SR 125\n(Roadway Minor Rehab)"
)

case_colors <- c(
  "2024_000000000018030" = COL_GREEN,
  "2024_000000000271525" = COL_RED,
  "2024_000000000270710" = COL_MUTED
)

# Pull yearly crash rates for each case study segment
case_yearly <- segment_year_data %>%
  filter(ROADWAY_INVENTORY_ID %in% case_ids) %>%
  left_join(
    case_study_candidates %>%
      filter(ROADWAY_INVENTORY_ID %in% case_ids) %>%
      distinct(ROADWAY_INVENTORY_ID, completion_year, PROJECT_NME,
               PRIMARY_WORK_CATEGORY, EST_TOTAL_CONSTR_COST,
               before_rate, after_rate, change),
    by = "ROADWAY_INVENTORY_ID"
  ) %>%
  mutate(
    label     = case_labels[ROADWAY_INVENTORY_ID],
    seg_color = case_colors[ROADWAY_INVENTORY_ID]
  )


case_yearly <- case_yearly %>%
  left_join(
    analysis_df %>%
      st_drop_geometry() %>%
      select(ROADWAY_INVENTORY_ID, FUNCTION_CLASS_CD) %>%
      rename(FC = FUNCTION_CLASS_CD),
    by = "ROADWAY_INVENTORY_ID"
  ) %>%
  mutate(FUNCTION_CLASS_CD = FC) %>%
  select(-FC)

# Check we have data
case_yearly %>%
  group_by(ROADWAY_INVENTORY_ID, label) %>%
  summarise(
    years     = paste(sort(unique(CRASH_YR)), collapse = ", "),
    min_rate  = round(min(crash_rate_MVMT, na.rm = TRUE), 2),
    max_rate  = round(max(crash_rate_MVMT, na.rm = TRUE), 2),
    .groups   = "drop"
  ) %>%
  as.data.frame()
```

```{r}
############################################
### Case Study Visualizations
############################################

# ── Individual yearly trend plots ─────────────────────────────────────────────

plot_case_study <- function(seg_id) {
  
  seg_data <- case_yearly %>%
    filter(ROADWAY_INVENTORY_ID == seg_id)
  
  comp_year  <- unique(seg_data$completion_year)
  proj_name  <- unique(seg_data$PROJECT_NME)
  work_cat   <- unique(seg_data$PRIMARY_WORK_CATEGORY)
  cost       <- scales::dollar(unique(seg_data$EST_TOTAL_CONSTR_COST))
  before     <- round(unique(seg_data$before_rate), 1)
  after      <- round(unique(seg_data$after_rate), 1)
  chg        <- round(unique(seg_data$change), 1)
  seg_color  <- unique(seg_data$seg_color)
  
  chg_label  <- ifelse(chg < 0,
                       paste0("▼ ", abs(chg), " per MVMT"),
                       paste0("▲ ", abs(chg), " per MVMT"))
  chg_color  <- ifelse(chg < 0, COL_GREEN, COL_RED)
  
  ggplot(seg_data, aes(x = CRASH_YR, y = crash_rate_MVMT)) +
    geom_area(fill = seg_color, alpha = 0.1) +
    geom_line(color = seg_color, linewidth = 1.3) +
    geom_point(color = seg_color, size = 4) +
    geom_vline(xintercept = comp_year, linetype = "dashed",
               color = COL_ORANGE, linewidth = 0.8) +
    annotate("text", x = comp_year + 0.08,
             y = max(seg_data$crash_rate_MVMT, na.rm = TRUE) * 0.95,
             label = paste0("Completed\n", comp_year),
             color = COL_ORANGE, hjust = 0, size = 5.5) +
    annotate("label",
             x = mean(range(seg_data$CRASH_YR)),
             y = max(seg_data$crash_rate_MVMT, na.rm = TRUE) * 0.5,
             label = paste0("Before: ", before,
                            "\nAfter:  ", after,
                            "\nChange: ", chg_label),
             fill = "#1c2128", color = chg_color,
             size = 5.5, label.size = 0.5,
             label.padding = unit(0.4, "lines")) +
    scale_x_continuous(breaks = unique(seg_data$CRASH_YR)) +
    labs(
      title    = proj_name,
      subtitle = paste0(work_cat, " | Cost: ", cost),
      x        = "Year",
      y        = "Crash Rate (per MVMT)",
      caption  = paste0("Segment ID: ", seg_id,
                        " | FC: ", unique(seg_data$FUNCTION_CLASS_CD),
                        " | ADT: ", 
                        scales::comma(unique(seg_data$ADT_TOTAL_NBR)))
    ) +
    theme_odot()
}

# Generate all three plots
p_intersection <- plot_case_study("2024_000000000018030")
p_traffic      <- plot_case_study("2024_000000000271525")
p_rehab        <- plot_case_study("2024_000000000270710")

p_intersection
p_traffic
p_rehab

# ── Summary comparison panel ──────────────────────────────────────────────────

summary_data <- case_study_candidates %>%
  filter(ROADWAY_INVENTORY_ID %in% case_ids) %>%
  distinct(ROADWAY_INVENTORY_ID, .keep_all = TRUE) %>%
  mutate(
    label     = case_labels[ROADWAY_INVENTORY_ID],
    seg_color = case_colors[ROADWAY_INVENTORY_ID],
    outcome   = case_when(
      change < 0 ~ "Improved",
      change > 0 ~ "Worsened",
      TRUE       ~ "No Change"
    )
  ) %>%
  pivot_longer(cols = c(before_rate, after_rate),
               names_to = "period", values_to = "rate") %>%
  mutate(period = factor(ifelse(period == "before_rate", 
                                "Before", "After"),
                         levels = c("Before", "After")))

p_summary <- ggplot(summary_data,
                    aes(x = period, y = rate,
                        color = ROADWAY_INVENTORY_ID,
                        group = ROADWAY_INVENTORY_ID)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 4) +
  geom_label(aes(label = round(rate, 1)),
             nudge_x = 0.08, show.legend = FALSE,
             fill = "#161b22", size = 5.5, linewidth = 0) +
  scale_color_manual(values = case_colors,
                     labels = c(
                       "2024_000000000018030" = "Plainfield Roundabouts",
                       "2024_000000000271525" = "Eastern Corridor Traffic Control",
                       "2024_000000000270710" = "SR 32/SR 125 Rehab"
                     )) +
  labs(
    title    = "Case Study Comparison: Three Hamilton County Projects",
    subtitle = "Before vs after crash rates | Each line is one project",
    x        = NULL,
    y        = "Avg Crash Rate (per MVMT)",
    color    = NULL,
    caption  = "Hamilton County, Ohio | Safety-funded projects completing 2021–2023"
  ) +
  theme_odot() +
  theme(legend.position = "bottom")

p_summary

# ── Combined panel ─────────────────────────────────────────────────────────────
case_study_panel <- (p_intersection | p_traffic | p_rehab) / p_summary +
  plot_annotation(
    title    = "Hamilton County Case Studies: What Do Safety Projects Actually Do?",
    subtitle = "Three projects — three different outcomes",
    caption  = "Source: ODOT Construction Projects & Crash Data 2020–2024",
    theme    = theme(
      plot.background = element_rect(fill = "#0d1117", color = NA),
      plot.title      = element_text(color = "#e6edf3", size = 16,
                                     face = "bold"),
      plot.subtitle   = element_text(color = "#8b949e", size = 11),
      plot.caption    = element_text(color = "#8b949e", size = 9)
    )
  )

case_study_panel

#ggsave("hamilton_case_studies.png", case_study_panel,
 #      width = 16, height = 10, dpi = 150, bg = "#0d1117")


# Build from scratch using ALL project types
hamilton_all_types <- analysis_df %>%
  st_drop_geometry() %>%
  filter(COUNTY_CD.x == "HAM") %>%
  inner_join(
    projects_safety %>%
      select(NLF_ID, CTL_BEGIN, CTL_END, completion_year,
             PRIMARY_WORK_CATEGORY, EST_TOTAL_CONSTR_COST),
    by = "NLF_ID",
    relationship = "many-to-many"
  ) %>%
  filter(
    CTL_BEGIN_NBR <= CTL_END.y,
    CTL_END_NBR   >= CTL_BEGIN.y
  ) %>%
  distinct(ROADWAY_INVENTORY_ID, PRIMARY_WORK_CATEGORY, 
           completion_year, .keep_all = TRUE)

# Check all categories present
cat("Categories found:", n_distinct(hamilton_all_types$PRIMARY_WORK_CATEGORY), "\n")

# Join yearly data and summarise
segment_year_data %>%
  filter(ROADWAY_INVENTORY_ID %in% 
           hamilton_all_types$ROADWAY_INVENTORY_ID) %>%
  left_join(
    hamilton_all_types %>%
      select(ROADWAY_INVENTORY_ID, PRIMARY_WORK_CATEGORY),
    by = "ROADWAY_INVENTORY_ID"
  ) %>%
  group_by(PRIMARY_WORK_CATEGORY) %>%
  summarise(
    n_segments = n_distinct(ROADWAY_INVENTORY_ID),
    avg_before = round(mean(crash_rate_MVMT[period == "before"], 
                            na.rm = TRUE), 2),
    avg_after  = round(mean(crash_rate_MVMT[period == "after"],  
                            na.rm = TRUE), 2),
    .groups    = "drop"
  ) %>%
  mutate(
    change    = round(avg_after - avg_before, 3),
    direction = case_when(
      change < -0.5 ~ "Improved",
      change >  0.5 ~ "Worsened",
      TRUE          ~ "No Change"
    )
  ) %>%
  arrange(change) %>%
  as.data.frame()
```

```{r}
# Cost per unit of crash reduction
case_study_candidates %>%
  filter(ROADWAY_INVENTORY_ID %in% case_ids) %>%
  distinct(ROADWAY_INVENTORY_ID, .keep_all = TRUE) %>%
  mutate(
    # Cost per crash rate point reduced
    cost_per_unit_reduction = EST_TOTAL_CONSTR_COST / abs(change),
    # Annualized cost effectiveness
    crashes_reduced_per_million = (abs(change) / 
                                  EST_TOTAL_CONSTR_COST) * 1e6
  ) %>%
  select(PROJECT_NME, PRIMARY_WORK_CATEGORY,
         EST_TOTAL_CONSTR_COST, change,
         cost_per_unit_reduction,
         crashes_reduced_per_million) %>%
  as.data.frame()
```

```{r}
# Add cost effectiveness to the summary comparison plot
cost_data <- data.frame(
  project = c("Plainfield\nRoundabouts",
              "Eastern Corridor\nTraffic Control",
              "SR 32/SR 125\nRehab"),
  cost_millions     = c(8.863, 0.337, 1.616),
  crashes_per_million = c(5.10, -44.34, 0.00),
  color = c(COL_GREEN, COL_RED, COL_MUTED)
)

p_cost <- ggplot(cost_data,
                 aes(x = reorder(project, crashes_per_million),
                     y = crashes_per_million,
                     fill = color)) +
  geom_col(width = 0.55, alpha = 0.85) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = COL_MUTED, linewidth = 0.7) +
  geom_text(aes(label = paste0(crashes_per_million,
                               "\nper $1M")),
            vjust = ifelse(cost_data$crashes_per_million >= 0,
                          -0.3, 1.3),
            color = "#e6edf3", size = 5.5) +
  scale_fill_identity() +
  coord_flip() +
  labs(
    title    = "Cost Effectiveness: Crashes Reduced per $1M Spent",
    subtitle = "Hamilton County case studies | Negative = crashes increased",
    x        = NULL,
    y        = "Crash Rate Units Reduced per $1M",
    caption  = "Positive = safety improvement per dollar | Negative = safety worsened per dollar"
  ) +
  theme_odot()

p_cost
```

```{r}
# Find more candidates excluding all projects we've already used
used_ids <- c(
  "2024_000000000018030",  # Plainfield Roundabouts
  "2024_000000000271525",  # Eastern Corridor VAR TSG
  "2024_000000000270710",  # SR 32/SR 125
  "2024_000000000270927",  # Cincinnati PSIP Pedestrian
  "2024_000000000119861",  # GR Type A Roadside
  "2024_000000000120118"   # Cincinnati Friction
)

new_candidates_3 <- hamilton_all_types %>%
  filter(!ROADWAY_INVENTORY_ID %in% used_ids) %>%
  left_join(
    projects_safety %>%
      select(NLF_ID, PROJECT_NME) %>%
      distinct(),
    by = "NLF_ID"
  ) %>%
  left_join(
    segment_year_data %>%
      group_by(ROADWAY_INVENTORY_ID) %>%
      summarise(
        before_rate = mean(crash_rate_MVMT[period == "before"], na.rm = TRUE),
        after_rate  = mean(crash_rate_MVMT[period == "after"],  na.rm = TRUE),
        change      = after_rate - before_rate,
        .groups     = "drop"
      ),
    by = "ROADWAY_INVENTORY_ID"
  ) %>%
  filter(!is.na(change)) %>%
  distinct(ROADWAY_INVENTORY_ID, PRIMARY_WORK_CATEGORY, .keep_all = TRUE) %>%
  select(ROADWAY_INVENTORY_ID, PROJECT_NME, PRIMARY_WORK_CATEGORY,
         COUNTY_CD.x, completion_year, before_rate, after_rate,
         change, EST_TOTAL_CONSTR_COST, FUNCTION_CLASS_CD,
         ADT_TOTAL_NBR) %>%
  arrange(change)

as.data.frame(new_candidates_3)
```

```{r}


```

```{r}


```

```{r}
############################################
### All Ohio Projects — Trend Per Category
### Centered on event time (t=0 = completion)
############################################

# Build statewide version with all categories
ohio_all_types <- analysis_df %>%
  st_drop_geometry() %>%
  inner_join(
    projects_safety %>%
      select(NLF_ID, CTL_BEGIN, CTL_END, completion_year,
             PRIMARY_WORK_CATEGORY, EST_TOTAL_CONSTR_COST),
    by = "NLF_ID",
    relationship = "many-to-many"
  ) %>%
  filter(
    CTL_BEGIN_NBR <= CTL_END.y,
    CTL_END_NBR   >= CTL_BEGIN.y,
    !is.na(completion_year)
  ) %>%
  distinct(ROADWAY_INVENTORY_ID, PRIMARY_WORK_CATEGORY,
           completion_year, .keep_all = TRUE)

# Join yearly crash rates and compute event time
ohio_yearly_full <- segment_year_data %>%
  filter(ROADWAY_INVENTORY_ID %in%
           ohio_all_types$ROADWAY_INVENTORY_ID) %>%
  left_join(
    ohio_all_types %>%
      select(ROADWAY_INVENTORY_ID, PRIMARY_WORK_CATEGORY,
             completion_year) %>%
      distinct(ROADWAY_INVENTORY_ID, PRIMARY_WORK_CATEGORY,
               completion_year),
    by = "ROADWAY_INVENTORY_ID",
    relationship = "many-to-many"
  ) %>%
  filter(
    !is.na(PRIMARY_WORK_CATEGORY),
    !is.na(completion_year.y)        # <-- use .y suffix
  ) %>%
  mutate(
    event_time = CRASH_YR - completion_year.y
  ) %>%
  filter(event_time >= -2, event_time <= 2)

# Check coverage
ohio_yearly_full %>%
  group_by(PRIMARY_WORK_CATEGORY) %>%
  summarise(
    n_segments  = n_distinct(ROADWAY_INVENTORY_ID),
    years_avail = paste(sort(unique(event_time)), collapse = ", "),
    .groups     = "drop"
  ) %>%
  arrange(desc(n_segments)) %>%
  as.data.frame()


############################################
### Ohio Statewide Event Study — All Categories
############################################

# Summarise by category and event time
ohio_event_summary <- ohio_yearly_full %>%
  group_by(PRIMARY_WORK_CATEGORY, event_time) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    se             = sd(crash_rate_MVMT, na.rm = TRUE) / sqrt(n()),
    n_segments     = n_distinct(ROADWAY_INVENTORY_ID),
    .groups        = "drop"
  ) %>%
  group_by(PRIMARY_WORK_CATEGORY) %>%
  mutate(
    did_change = mean(avg_crash_rate[event_time >= 0], na.rm = TRUE) -
                 mean(avg_crash_rate[event_time < 0],  na.rm = TRUE),
    direction  = case_when(
      did_change < -0.5 ~ "Improved",
      did_change >  0.5 ~ "Worsened",
      TRUE              ~ "No Change"
    )
  ) %>%
  ungroup()

# Color by outcome
outcome_colors <- c(
  "Improved"  = COL_GREEN,
  "Worsened"  = COL_RED,
  "No Change" = COL_MUTED
)

# Build one plot per category
ohio_plots <- lapply(
  unique(ohio_event_summary$PRIMARY_WORK_CATEGORY),
  function(cat) {
    
    cat_data   <- ohio_event_summary %>% 
                    filter(PRIMARY_WORK_CATEGORY == cat)
    direction  <- unique(cat_data$direction)
    seg_color  <- outcome_colors[direction]
    n_segs     <- max(cat_data$n_segments)
    chg        <- round(unique(cat_data$did_change), 2)
    chg_label  <- ifelse(chg < 0,
                         paste0("▼ ", abs(chg), "/MVMT"),
                         paste0("▲ ", abs(chg), "/MVMT"))
    sparse     <- n_segs < 20
    
    ggplot(cat_data,
           aes(x = event_time, y = avg_crash_rate)) +
      geom_ribbon(aes(ymin = avg_crash_rate - 1.96 * se,
                      ymax = avg_crash_rate + 1.96 * se),
                  fill = seg_color, alpha = 0.15) +
      geom_line(color = seg_color, linewidth = 1.2) +
      geom_point(color = seg_color, size = 3) +
      geom_vline(xintercept = 0, linetype = "dashed",
                 color = COL_ORANGE, linewidth = 0.7) +
      annotate("text", x = 0.1,
               y = max(cat_data$avg_crash_rate +
                         1.96 * cat_data$se, na.rm = TRUE) * 0.97,
               label = "Completion",
               color = COL_ORANGE, hjust = 0, size = 5.0) +
      annotate("label",
               x = -1.9,
               y = max(cat_data$avg_crash_rate +
                         1.96 * cat_data$se, na.rm = TRUE) * 0.6,
               label = paste0(direction, ": ", chg_label,
                              if (sparse) "\n⚠ Low n" else ""),
               fill    = "#1c2128",
               color   = seg_color,
               size    = 2.8,
               label.size     = 0.4,
               label.padding  = unit(0.3, "lines"),
               hjust   = 0) +
      scale_x_continuous(breaks = -2:2,
                         labels = c("-2", "-1", "0\n(Completion)",
                                    "+1", "+2")) +
      labs(
        title    = cat,
        subtitle = paste0("n = ", n_segs,
                          " segments | Ohio Statewide | 95% CI shown"),
        x        = "Years Relative to Project Completion",
        y        = "Avg Crash Rate (per MVMT)"
      ) +
      theme_odot() +
      theme(plot.title = element_text(size = 11))
  }
)

for (p in ohio_plots) { print(p) }
```

```{r}
#####################
### updated fig 6
#####################

# Rebuild fig6 investment gap data with funding % instead of segment %
fc_labels <- c(
  "1" = "FC1: Interstate",
  "2" = "FC2: Freeway",
  "3" = "FC3: Major Arterial",
  "4" = "FC4: Minor Arterial",
  "5" = "FC5: Collector",
  "6" = "FC6: Minor Collector",
  "7" = "FC7: Local"
)

investment_gap_funding <- data.frame(
  FUNCTION_CLASS_CD = c(1,2,3,4,5,6,7),
  pct_funding       = c(22.0, 6.9, 39.5, 19.8, 9.8, 0.6, 1.4),
  crash_rate        = c(11.3, 8.0, 23.7, 25.6, 25.0, 22.6, 36.5),
  pct_fatalities    = c(8.1, 2.3, 19.3, 24.0, 28.5, 4.9, 12.8)
) %>%
  mutate(
    fc_label   = fc_labels[as.character(FUNCTION_CLASS_CD)],
    bubble_size = pct_fatalities
  )

fig6_updated <- ggplot(investment_gap_funding,
       aes(x = pct_funding,
           y = crash_rate,
           size = bubble_size,
           color = factor(FUNCTION_CLASS_CD))) +
  geom_point(alpha = 0.85) +
  geom_text(aes(label = fc_label),
            vjust = -1.2, size = 4.0,
            color = "#e6edf3", fontface = "bold") +
  annotate("rect", xmin = 0, xmax = 12,
           ymin = 22, ymax = 40,
           fill = COL_RED, alpha = 0.06) +
  annotate("text", x = 6, y = 39,
           label = "High risk · Low funding",
           color = COL_RED, size = 3, fontface = "italic") +
  annotate("rect", xmin = 25, xmax = 70,
           ymin = 0, ymax = 15,
           fill = COL_BLUE, alpha = 0.06) +
  annotate("text", x = 47, y = 14,
           label = "Low risk · High funding",
           color = COL_BLUE, size = 4, fontface = "italic") +
  scale_size_continuous(range = c(4, 18),
                        name = "% of Fatalities") +
  scale_color_manual(values = c(
    "1" = COL_BLUE, "2" = COL_BLUE,
    "3" = COL_YELLOW, "4" = COL_ORANGE,
    "5" = COL_RED, "6" = COL_RED,
    "7" = COL_RED
  ), guide = "none") +
  labs(
    title    = "Safety Investment Gap by Road Class",
    subtitle = "Bubble size = share of statewide fatalities",
    x        = "% of Total Safety Funding Received",
    y        = "Avg Crash Rate (per MVMT)",
    caption  = "Source: ODOT Construction Projects & Crash Data 2020–2024"
  ) +
  theme_odot()

fig6_updated
```

```{r}
# Convert 2022 crashes to sf and spatial join to segments
crashes_2022_sf <- Crash2022 %>%
  filter(!is.na(ODOT_LATITUDE_NBR), !is.na(ODOT_LONGITUDE_NBR)) %>%
  st_as_sf(coords = c("ODOT_LONGITUDE_NBR", "ODOT_LATITUDE_NBR"),
           crs = 4326) %>%
  st_transform(st_crs(analysis_df))

# Spatial join to nearest segment
crashes_2022_joined <- crashes_2022_sf %>%
  mutate(
    ROADWAY_INVENTORY_ID = analysis_df$ROADWAY_INVENTORY_ID[
      st_nearest_feature(., analysis_df)
    ]
  ) %>%
  st_drop_geometry()

# Aggregate to segment level
crashes_2022_summary <- crashes_2022_joined %>%
  group_by(ROADWAY_INVENTORY_ID) %>%
  summarise(
    yearly_crashes = n(),
    yearly_fatal   = sum(CRASH_SEVERITY_CD == 1, na.rm = TRUE),
    yearly_injury  = sum(CRASH_SEVERITY_CD %in% c(2,3), na.rm = TRUE),
    CRASH_YR       = 2022,
    .groups        = "drop"
  ) %>%
  left_join(
    analysis_df %>%
      st_drop_geometry() %>%
      select(ROADWAY_INVENTORY_ID, SEGMENT_LENGTH_NBR, ADT_TOTAL_NBR),
    by = "ROADWAY_INVENTORY_ID"
  ) %>%
  mutate(
    MVMT           = ADT_TOTAL_NBR * SEGMENT_LENGTH_NBR * 365 / 1e6,
    crash_rate_MVMT = ifelse(MVMT > 0, yearly_crashes / MVMT, NA),
    period         = "transition",
    treated_safety = ROADWAY_INVENTORY_ID %in% 
                       treated_safety_ids$ROADWAY_INVENTORY_ID,
    completion_year = NA_real_,
    event_time     = NA_real_
  )

cat("2022 segments processed:", nrow(crashes_2022_summary), "\n")
cat("2022 treated segments:", 
    sum(crashes_2022_summary$treated_safety), "\n")

# Add 2022 to segment_year_data
crashes_2022_summary <- crashes_2022_summary %>%
  mutate(
    treated        = ROADWAY_INVENTORY_ID %in% 
                       analysis_df$ROADWAY_INVENTORY_ID,
    treated_safety = ROADWAY_INVENTORY_ID %in%
                       treated_safety_ids$ROADWAY_INVENTORY_ID,
    period         = "transition",
    completion_year = NA_real_,
    event_time     = NA_real_
  )

# Stack with existing data
segment_year_data_full <- bind_rows(
  segment_year_data,
  crashes_2022_summary %>%
    select(any_of(names(segment_year_data)))
) %>%
  arrange(ROADWAY_INVENTORY_ID, CRASH_YR)

# Recompute event time for treated segments
segment_year_data_full <- segment_year_data_full %>%
  left_join(
    ohio_all_types %>%
      distinct(ROADWAY_INVENTORY_ID, completion_year) %>%
      rename(comp_year = completion_year),
    by = "ROADWAY_INVENTORY_ID"
  ) %>%
  mutate(
    event_time = CRASH_YR - comp_year
  )

# Rebuild event study with 2022 included
event_study_full <- segment_year_data_full %>%
  filter(
    treated_safety == TRUE,
    !is.na(event_time),
    event_time >= -2, event_time <= 2
  ) %>%
  group_by(event_time) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    se             = sd(crash_rate_MVMT,   na.rm = TRUE) / sqrt(n()),
    n_obs          = n(),
    .groups        = "drop"
  )

as.data.frame(event_study_full)
```

```{r}

completion_lookup <- analysis_df %>%
  st_drop_geometry() %>%
  select(ROADWAY_INVENTORY_ID, NLF_ID, CTL_BEGIN_NBR, CTL_END_NBR) %>%
  inner_join(
    projects_safety %>%
      select(NLF_ID, CTL_BEGIN, CTL_END, completion_year) %>%
      distinct(),
    by = "NLF_ID",
    relationship = "many-to-many"
  ) %>%
  filter(
    CTL_BEGIN_NBR <= CTL_END,   # segment start <= project end
    CTL_END_NBR   >= CTL_BEGIN  # segment end >= project start
  ) %>%
  group_by(ROADWAY_INVENTORY_ID) %>%
  summarise(comp_year = min(completion_year, na.rm = TRUE),
            .groups = "drop")

# Recompute event time
segment_year_data_full <- segment_year_data_full %>%
  left_join(completion_lookup, by = "ROADWAY_INVENTORY_ID") %>%
  mutate(event_time_full = CRASH_YR - comp_year)

# Rebuild event study with 2022 included
event_study_full <- segment_year_data_full %>%
  filter(
    treated_safety == TRUE,
    !is.na(event_time_full),
    event_time_full >= -2, event_time_full <= 2
  ) %>%
  group_by(event_time_full) %>%
  summarise(
    avg_crash_rate = mean(crash_rate_MVMT, na.rm = TRUE),
    se             = sd(crash_rate_MVMT,   na.rm = TRUE) / sqrt(n()),
    n_obs          = n(),
    .groups        = "drop"
  )

cat("=== Event Study WITH 2022 ===\n")
as.data.frame(event_study_full)

cat("\n=== Original Event Study (without 2022) ===\n")
data.frame(
  event_time     = c(-2, -1, 0, 1, 2),
  avg_crash_rate = c(8.10, 7.64, 6.74, 6.80, 7.42)
)
```

```{r}
#####################
### Testing Clusters
#####################

# Step 1 — create balanced_segments_nlf
balanced_segments_nlf <- balanced_segments %>%
  left_join(
    analysis_df %>%
      st_drop_geometry() %>%
      select(ROADWAY_INVENTORY_ID, NLF_ID, COUNTY_CD.x),
    by = "ROADWAY_INVENTORY_ID"
  )

# Step 2 — all three models
did_segment <- felm(log_crash_rate ~ treated_safety * I(period == "after") |
                      ROADWAY_INVENTORY_ID + CRASH_YR | 0 |
                      ROADWAY_INVENTORY_ID,
                    data = balanced_segments_nlf %>%
                      filter(is.finite(log_crash_rate)))

did_nlf <- felm(log_crash_rate ~ treated_safety * I(period == "after") |
                  ROADWAY_INVENTORY_ID + CRASH_YR | 0 |
                  NLF_ID,
                data = balanced_segments_nlf %>%
                  filter(is.finite(log_crash_rate)))

did_county <- felm(log_crash_rate ~ treated_safety * I(period == "after") |
                     ROADWAY_INVENTORY_ID + CRASH_YR | 0 |
                     COUNTY_CD.x,
                   data = balanced_segments_nlf %>%
                     filter(is.finite(log_crash_rate)))

# Step 3 — compare
data.frame(
  clustering = c("Segment", "NLF_ID", "County"),
  estimate   = c(
    coef(did_segment)["treated_safetyTRUE:I(period == \"after\")TRUE"],
    coef(did_nlf)["treated_safetyTRUE:I(period == \"after\")TRUE"],
    coef(did_county)["treated_safetyTRUE:I(period == \"after\")TRUE"]
  ),
  se         = c(
    did_segment$cse["treated_safetyTRUE:I(period == \"after\")TRUE"],
    did_nlf$cse["treated_safetyTRUE:I(period == \"after\")TRUE"],
    did_county$cse["treated_safetyTRUE:I(period == \"after\")TRUE"]
  )
) %>%
  mutate(p_value = round(2 * pnorm(-abs(estimate / se)), 3))
```

```{r}

# Rebuild treated_flag
treated_flag <- segment_year_data %>%
  st_drop_geometry() %>%
  distinct(ROADWAY_INVENTORY_ID, treated_safety)

# Rebuild analysis_app
analysis_app <- analysis_df %>%
  st_drop_geometry() %>%
  select(
    ROADWAY_INVENTORY_ID, NLF_ID, COUNTY_CD.x,
    FUNCTION_CLASS_CD, ADT_TOTAL_NBR, SPEED_LIMIT_NBR,
    SEGMENT_LENGTH_NBR, LANES_NBR, PCR_NBR, AVG_IRI,
    total_crashes, fatal_crashes, injury_crashes,
    crashes_per_MVMT, treated
  ) %>%
  left_join(treated_flag, by = "ROADWAY_INVENTORY_ID")

# Rebuild segment_year_app
segment_year_app <- segment_year_data %>%
  st_drop_geometry() %>%
  select(
    ROADWAY_INVENTORY_ID, CRASH_YR, period,
    yearly_crashes, yearly_fatal, yearly_injury,
    crash_rate_MVMT, treated_safety, completion_year,
    ADT_TOTAL_NBR, SEGMENT_LENGTH_NBR, MVMT
  )

# Rebuild projects_app
projects_app <- projects_safety %>%
  select(
    NLF_ID, CTL_BEGIN, CTL_END, completion_year,
    PRIMARY_WORK_CATEGORY, EST_TOTAL_CONSTR_COST,
    PROJECT_NME, COUNTY_NME, DISTRICT_NBR,
    SAFETY_FUNDS_IND
  ) %>%
  distinct()

# Rebuild county_summary
county_summary <- analysis_app %>%
  left_join(
    projects_app %>%
      select(NLF_ID, EST_TOTAL_CONSTR_COST) %>%
      group_by(NLF_ID) %>%
      summarise(total_cost = sum(EST_TOTAL_CONSTR_COST,
                                 na.rm = TRUE),
                .groups = "drop"),
    by = "NLF_ID"
  ) %>%
  group_by(COUNTY_CD.x, FUNCTION_CLASS_CD) %>%
  summarise(
    n_segments       = n(),
    n_treated        = sum(treated_safety, na.rm = TRUE),
    total_crashes    = sum(total_crashes, na.rm = TRUE),
    total_fatalities = sum(fatal_crashes, na.rm = TRUE),
    avg_crash_rate   = mean(crashes_per_MVMT, na.rm = TRUE),
    total_funding    = sum(total_cost, na.rm = TRUE),
    .groups          = "drop"
  ) %>%
  group_by(COUNTY_CD.x) %>%
  mutate(
    pct_funding    = round(100 * total_funding /
                             sum(total_funding, na.rm = TRUE), 1),
    pct_fatalities = round(100 * total_fatalities /
                             sum(total_fatalities, na.rm = TRUE), 1),
    pct_segments   = round(100 * n_treated / n_segments, 1)
  ) %>%
  ungroup()

# Save everything
saveRDS(analysis_app,     "app_data/analysis_app.rds")
saveRDS(segment_year_app, "app_data/segment_year_app.rds")
saveRDS(projects_app,     "app_data/projects_app.rds")
saveRDS(county_summary,   "app_data/county_summary.rds")

message("All app data saved successfully")
```

### Making the App

```{r}
# Create app folder structure
dir.create("odot_app", showWarnings = FALSE)
dir.create("odot_app/app_data", showWarnings = FALSE)

# Copy all data files into app folder
file.copy(list.files("app_data", full.names = TRUE),
          "odot_app/app_data/",
          overwrite = TRUE)

message("App data copied to odot_app/app_data/")

```

Current issues:

not displaying the right info for the projects - fixed

can't see if there is a certain type of project for a county before selecting it - fixed

needs an option for all counties - fixed

map controls don't seem to work and counties are not very clear (map kinda sucks in its current form...) - just decided to nuke it

need to make the county dashboard display on the page better - I understand the problem

need road names with the segment, or some way to view the segments on a map - yes kinda fixed\

move the map for the risk explorer to right under the filter section

the ADT sort is not working right.

remove really short segments with ungodly high crash rates from the risk analysis tab? maybe make a box you can check to turn them on or off.

find a substitution for the overview map only if it can add a level of significant important or be interesting/cool.

get started on presentation. build a background, show results, create a story, and use the app to demonstrate it's use.

```{r}

```

```{r}
############################################
### ODOT Safety Investment Analyzer
### Shiny App — app.R
############################################

library(shiny)
library(shinydashboard)
library(tidyverse)
library(sf)
library(leaflet)
library(patchwork)
library(scales)
library(ggplot2)
library(DT)

# ── Load data ─────────────────────────────────────────────────────────────────
analysis_app             <- readRDS("app_data/analysis_app.rds")
segment_year_app         <- readRDS("app_data/segment_year_app.rds")
projects_app             <- readRDS("app_data/projects_app.rds")
county_summary           <- readRDS("app_data/county_summary.rds")
segments_geo             <- readRDS("app_data/segments_geo.rds")
project_segment_lookup   <- readRDS("app_data/project_segment_lookup.rds")

# ── County name lookup ────────────────────────────────────────────────────────
county_name_lookup <- analysis_app %>%
  left_join(
    projects_app %>%
      select(COUNTY_NME) %>%
      distinct() %>%
      mutate(COUNTY_CD.x = toupper(substr(COUNTY_NME, 1, 3))),
    by = "COUNTY_CD.x"
  ) %>%
  distinct(COUNTY_CD.x, COUNTY_NME) %>%
  filter(!is.na(COUNTY_NME)) %>%
  mutate(display = paste0(COUNTY_CD.x, " — ", COUNTY_NME)) %>%
  arrange(COUNTY_CD.x)

county_choices <- setNames(
  c("ALL", county_name_lookup$COUNTY_CD.x),
  c("Statewide — All Ohio", county_name_lookup$display)
)

# ── Constants ─────────────────────────────────────────────────────────────────
COL_ORANGE <- "#f0883e"
COL_BLUE   <- "#58a6ff"
COL_GREEN  <- "#3fb950"
COL_RED    <- "#f85149"
COL_YELLOW <- "#d29922"
COL_MUTED  <- "#8b949e"

##county_choices <- sort(unique(analysis_app$COUNTY_CD.x))

fc_labels <- c(
  "1" = "FC1: Interstate",
  "2" = "FC2: Freeway",
  "3" = "FC3: Major Arterial",
  "4" = "FC4: Minor Arterial",
  "5" = "FC5: Collector",
  "6" = "FC6: Minor Collector",
  "7" = "FC7: Local"
)

theme_odot <- function(base_size = 16) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background   = element_rect(fill = "#F0F4F8", color = NA),
      panel.background  = element_rect(fill = "#FFFFFF", color = NA),
      panel.grid.major  = element_line(color = "#D1DCE8", linewidth = 0.4),
      panel.grid.minor  = element_blank(),
      text              = element_text(color = "#0D1B2A", size = base_size),
      axis.text         = element_text(color = "#1A2E45", size = base_size - 2),
      axis.title        = element_text(color = "#1A2E45", size = base_size - 1),
      plot.title        = element_text(color = "#0D1B2A", face = "bold",
                                       size = base_size + 4),
      plot.subtitle     = element_text(color = "#1A2E45", size = base_size),
      plot.caption      = element_text(color = "#7A9BB5", size = base_size - 3),
      legend.background = element_rect(fill = "#F0F4F8", color = NA),
      legend.text       = element_text(color = "#1A2E45", size = base_size - 2),
      legend.title      = element_text(color = "#1A2E45", size = base_size - 1),
      strip.text        = element_text(color = "#0D1B2A", size = base_size - 1)
    )
}

# ── Pre-build road name lookup from analysis_app ──────────────────────────────
road_name_lookup <- analysis_app %>%
  select(ROADWAY_INVENTORY_ID, NLF_ID) %>%
  left_join(
    projects_app %>%
      select(NLF_ID, PROJECT_NME) %>%
      distinct(),
    by = "NLF_ID"
  ) %>%
  distinct(ROADWAY_INVENTORY_ID, .keep_all = TRUE)

# ── Pre-build segment crash data lookup ───────────────────────────────────────
seg_crash_lookup <- segment_year_app %>%
  group_by(ROADWAY_INVENTORY_ID) %>%
  summarise(
    before_rate = mean(crash_rate_MVMT[period == "before"], na.rm = TRUE),
    after_rate  = mean(crash_rate_MVMT[period == "after"],  na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  mutate(has_data = !is.na(before_rate) | !is.na(after_rate))

############################################
### UI
############################################

ui <- dashboardPage(
  skin = "black",

  dashboardHeader(title = "ODOT Safety Analyzer"),

  dashboardSidebar(
    sidebarMenu(
      menuItem("County Dashboard",  tabName = "county",  icon = icon("chart-bar")),
      menuItem("Project Evaluator", tabName = "project", icon = icon("search")),
      menuItem("Risk Explorer",     tabName = "risk",    icon = icon("exclamation-triangle")),
      menuItem("Funding Gap", tabName = "funding", icon = icon("chart-pie")),
      menuItem("Cost Effectiveness", tabName = "cost", icon = icon("dollar-sign")),
      menuItem("Methodology",       tabName = "methods", icon = icon("info-circle"))
    )
  ),

  dashboardBody(
    tags$head(
      tags$style(HTML("
        .content-wrapper, .main-sidebar, .main-header .logo,
        .main-header .navbar { background-color: #0d1117 !important; }
        .skin-black .main-sidebar { background-color: #161b22 !important; }
        .skin-black .sidebar-menu > li.active > a,
        .skin-black .sidebar-menu > li:hover > a {
          border-left-color: #58a6ff !important; }
        .box { background: #161b22 !important;
               border-top-color: #58a6ff !important; }
        .box-header { color: #e6edf3 !important; }
        body { background-color: #0d1117 !important;
               color: #e6edf3 !important; }
        .selectize-input, .selectize-dropdown {
          background: #1c2128 !important;
          color: #e6edf3 !important; }
        table.dataTable thead th { color: #e6edf3 !important; }
        table.dataTable tbody td { color: #e6edf3 !important; }
        table.dataTable tbody tr { background-color: #161b22 !important; }
        table.dataTable tbody tr.odd  { background-color: #1c2128 !important; }
        table.dataTable tbody tr.even { background-color: #161b22 !important; }
        .dataTables_wrapper .dataTables_length label,
        .dataTables_wrapper .dataTables_filter label,
        .dataTables_wrapper .dataTables_info { color: #e6edf3 !important; }
        .dataTables_wrapper .dataTables_paginate .paginate_button {
          color: #e6edf3 !important; }
        .small-box { border-radius: 6px !important; }
        .value-box-text { font-size: 13px !important; }
      "))
    ),

    tabItems(

      # ── Tab 1: County Dashboard ────────────────────────────────────────────
      tabItem(tabName = "county",
        fluidRow(
          box(width = 3, title = "Select County",
            selectInput("county_select", "County:",
            choices  = county_choices,
            selected = "ALL"),
            hr(),
            valueBoxOutput("vbox_segments",   width = 12),
            valueBoxOutput("vbox_fatalities",  width = 12),
            valueBoxOutput("vbox_treated",     width = 12),
            hr(),
            p("Color guide:", style = "color:#8b949e; margin-bottom:4px;"),
            p("🔵 Interstate / Freeway", style = "color:#58a6ff; font-size:16px; margin:2px;"),
            p("🟡 Major Arterial",       style = "color:#d29922; font-size:16px; margin:2px;"),
            p("🔴 Minor Art. / Collector", style = "color:#f85149; font-size:16px; margin:2px;"),
            p("🟠 Minor Col. / Local",   style = "color:#f0883e; font-size:16px; margin:2px;")
          ),
          box(width = 9, title = textOutput("county_dashboard_title"),
            fluidRow(
              column(6, plotOutput("p_bar",     height = "380px")),
              column(6, plotOutput("p_fatal",   height = "380px"))
            ),
            fluidRow(
              column(6, plotOutput("p_funding", height = "380px")),
              column(6, plotOutput("p_bubble",  height = "380px"))
            )
          )
        )
      ),

      # ── Tab 2: Project Evaluator ───────────────────────────────────────────
      tabItem(tabName = "project",
  fluidRow(
    box(width = 3, title = "Select Project",
      radioButtons("proj_view", "View Mode:",
                   choices  = c("Individual Project" = "individual",
                                "Project Type Trend" = "aggregate"),
                   selected = "individual"),
      hr(),
      selectInput("proj_scope", "Geography:",
                  choices  = c("Statewide", "By County"),
                  selected = "Statewide"),
      conditionalPanel(
        condition = "input.proj_scope == 'By County'",
        selectInput("proj_county", "County:",
                    choices  = county_choices,
                    selected = "HAM")
      ),
      uiOutput("proj_category_ui"),
      conditionalPanel(
        condition = "input.proj_view == 'individual'",
        uiOutput("proj_name_ui")
      ),
      hr(),
      verbatimTextOutput("proj_details")
    ),
    box(width = 9, title = "Project Analysis",
      plotOutput("project_plot", height = "700px")
    )
  )
),

      # ── Tab 3: Risk Explorer ───────────────────────────────────────────────
       tabItem(tabName = "risk",
        fluidRow(
          box(width = 3, title = "Filters",
            selectInput("risk_county", "County:",
                        choices  = county_choices,
                        selected = "HAM"),
            selectInput("risk_fc", "Road Class:",
                        choices  = c("All", names(fc_labels)),
                        selected = "All"),
            sliderInput("risk_n", "Show top N segments:",
                        min = 10, max = 100, value = 25, step = 5),
            hr(),
            sliderInput("risk_min_length",
                        "Minimum segment length (miles):",
                        min = 0, max = 2, value = 0.1,
                        step = 0.05),
            checkboxInput("risk_exclude_short",
                          "Exclude very short segments (< min length)",
                          value = TRUE),
            hr(),
            p("Click any row to see segment on map.",
              style = "color:#8b949e; font-size:16px;"),
            leafletOutput("risk_map", height = "280px")
          ),
          box(width = 9, title = "Highest Risk Untreated Segments",
            DTOutput("risk_table")
          )
        )
      ),

      # ── Tab 4: Funding Gap ─────────────────────────────────────────────────
      tabItem(tabName = "funding",
        fluidRow(
          box(width = 3, title = "Filters",
            selectInput("fund_scope", "View:",
                        choices  = c("Statewide", "By County"),
                        selected = "Statewide"),
            conditionalPanel(
              condition = "input.fund_scope == 'By County'",
              selectInput("fund_county", "County:",
                          choices  = county_choices,
                          selected = "HAM")
            ),
            selectInput("fund_metric", "X Axis Metric:",
                        choices = c(
                          "% of Funding"          = "pct_funding",
                          "% of Segments Treated" = "pct_segments"
                        ))
          ),
          box(width = 9, title = "Investment Gap Analysis",
            plotOutput("funding_plot", height = "700px")
          )
        )
      ),


     # ── Tab 5: Cost Effectiveness ──────────────────────────────────────────
      tabItem(tabName = "cost",
        fluidRow(
          box(width = 3, title = "Filters",
            selectInput("cost_scope", "Geography:",
                        choices  = c("Statewide", "By County"),
                        selected = "Statewide"),
            conditionalPanel(
              condition = "input.cost_scope == 'By County'",
              selectInput("cost_county", "County:",
                          choices  = county_choices,
                          selected = "HAM")
            ),
            selectInput("cost_fc", "Road Class:",
                        choices = c("All", names(fc_labels)),
                        selected = "All"),
            hr(),
            p("Each bubble is one project.",
              style = "color:#8b949e; font-size:14px;"),
            p("X axis: total project cost.",
              style = "color:#8b949e; font-size:14px;"),
            p("Y axis: avg crash rate change after completion.",
              style = "color:#8b949e; font-size:14px;"),
            p("Bubble size: number of matched segments.",
              style = "color:#8b949e; font-size:14px;"),
            hr(),
            p("🟢 Meaningful improvement (> -0.5)",
              style = "color:#3fb950; font-size:14px;"),
            p("🔴 Meaningful worsening (> +0.5)",
              style = "color:#f85149; font-size:14px;"),
            p("⚫ Negligible change (within ±0.5)",
              style = "color:#8b949e; font-size:14px;")
          ),
          box(width = 9, title = "Cost vs Effectiveness",
            plotOutput("cost_plot", height = "650px")
          )
        )
      ),


      # ── Tab 6: Methodology ─────────────────────────────────────────────────
      tabItem(tabName = "methods",
        fluidRow(
          box(width = 12, title = "About This Analysis",
            h4("Data Sources", style = "color:#58a6ff"),
            p("This tool combines four ODOT datasets spanning 2020–2024:"),
            tags$ul(
              tags$li("Crash Records — ~910,000 events spatially joined to road segments"),
              tags$li("Road Inventory — 328,709 segments with physical characteristics"),
              tags$li("Pavement Condition (PCR) — most recent inspection per segment"),
              tags$li("Construction Projects — 13,550 safety-funded projects")
            ),
            hr(),
            h4("Analytical Approach", style = "color:#58a6ff"),
            p("The core analysis uses a difference-in-differences (DiD) design,
              comparing crash rate changes on safety-funded segments to comparable
              untreated segments over the same period. Treatment and control roads
              are matched by ADT range (10th–90th percentile of treated segments).
              Models include segment and year fixed effects with standard errors
              clustered at the segment level."),
            hr(),
            h4("Key Finding", style = "color:#58a6ff"),
            p("Safety-funded projects show no statistically significant overall
              crash reduction (DiD coefficient: +0.014, p = 0.185). Intersection
              improvements are the one project type showing consistent crash
              reductions (~6.5 crashes per MVMT). Local and collector roads
              account for over 50% of Ohio fatalities but receive less than 15%
              of safety funding."),
            hr(),
            p("Developed by Xavier University Senior Capstone, 2025–2026.",
              style = "color:#8b949e; font-style:italic")
          )
        )
      )
    )
  )
)

############################################
### SERVER
############################################

server <- function(input, output, session) {

  # ── Tab 1: County Dashboard ──────────────────────────────────────────────
   county_data <- reactive({
    if (input$county_select == "ALL") {
      county_summary %>%
        group_by(FUNCTION_CLASS_CD) %>%
        summarise(
          n_segments       = sum(n_segments),
          n_treated        = sum(n_treated),
          total_crashes    = sum(total_crashes),
          total_fatalities = sum(total_fatalities),
          avg_crash_rate   = mean(avg_crash_rate, na.rm = TRUE),
          total_funding    = sum(total_funding),
          .groups          = "drop"
        ) %>%
        mutate(
          pct_fatalities = round(100 * total_fatalities /
                                   sum(total_fatalities), 1),
          pct_funding    = round(100 * total_funding /
                                   sum(total_funding), 1),
          pct_segments   = round(100 * n_treated / n_segments, 1),
          COUNTY_CD.x    = "ALL",
          fc_label       = fc_labels[as.character(FUNCTION_CLASS_CD)],
          bar_color      = case_when(
            FUNCTION_CLASS_CD %in% c(1, 2) ~ COL_BLUE,
            FUNCTION_CLASS_CD == 3          ~ COL_YELLOW,
            FUNCTION_CLASS_CD %in% c(4, 5) ~ COL_RED,
            TRUE                            ~ COL_ORANGE
          )
        ) %>%
        filter(!is.na(fc_label))
    } else {
      county_summary %>%
        filter(COUNTY_CD.x == input$county_select) %>%
        mutate(
          fc_label  = fc_labels[as.character(FUNCTION_CLASS_CD)],
          bar_color = case_when(
            FUNCTION_CLASS_CD %in% c(1, 2) ~ COL_BLUE,
            FUNCTION_CLASS_CD == 3          ~ COL_YELLOW,
            FUNCTION_CLASS_CD %in% c(4, 5) ~ COL_RED,
            TRUE                            ~ COL_ORANGE
          )
        ) %>%
        filter(!is.na(fc_label))
    }
  })
   
    output$county_dashboard_title <- renderText({
    if (input$county_select == "ALL") {
      "Ohio Statewide — Safety Profile"
    } else {
      name <- county_name_lookup$COUNTY_NME[
        county_name_lookup$COUNTY_CD.x == input$county_select
      ]
      paste0(name, " County — Safety Profile")
    }
  })

  output$vbox_segments <- renderValueBox({
    d <- county_data()
    valueBox(
      value    = scales::comma(sum(d$n_segments)),
      subtitle = "Total Road Segments",
      icon     = icon("road"),
      color    = "blue",
      width    = 12
    )
  })

  output$vbox_fatalities <- renderValueBox({
    d <- county_data()
    valueBox(
      value    = scales::comma(sum(d$total_fatalities)),
      subtitle = "Fatalities 2020-2024",
      icon     = icon("exclamation-triangle"),
      color    = "red",
      width    = 12
    )
  })

  output$vbox_treated <- renderValueBox({
    d <- county_data()
    pct <- round(100 * sum(d$n_treated) / sum(d$n_segments), 1)
    valueBox(
      value    = paste0(pct, "%"),
      subtitle = "Segments w/ Investment",
      icon     = icon("wrench"),
      color    = "yellow",
      width    = 12
    )
  })

  # Individual plots for better layout control
  output$p_bar <- renderPlot({
    d <- county_data()
    req(nrow(d) > 0)
    ggplot(d, aes(x = reorder(fc_label, avg_crash_rate),
                  y = avg_crash_rate, fill = bar_color)) +
      geom_col(width = 0.6, alpha = 0.85) +
      geom_text(aes(label = round(avg_crash_rate, 1)),
                hjust = -0.2, color = "#0d1117", size = 5.5) +
      scale_fill_identity() +
      scale_y_continuous(limits = c(0, max(d$avg_crash_rate) * 1.3)) +
      coord_flip() +
      labs(title = "Crash Rate by Road Class",
           x = NULL, y = "Crashes per MVMT") +
      theme_odot()
  }, bg = "#0d1117")

  output$p_fatal <- renderPlot({
    d <- county_data()
    req(nrow(d) > 0)
    ggplot(d, aes(x = reorder(fc_label, pct_fatalities),
                  y = pct_fatalities, fill = bar_color)) +
      geom_col(width = 0.6, alpha = 0.85) +
      geom_text(aes(label = paste0(round(pct_fatalities, 1), "%")),
                hjust = -0.2, color = "#0d1117", size = 5.5) +
      scale_fill_identity() +
      scale_y_continuous(limits = c(0, max(d$pct_fatalities) * 1.3)) +
      coord_flip() +
      labs(title = "Share of Fatalities",
           x = NULL, y = "% of County Fatalities") +
      theme_odot()
  }, bg = "#0d1117")

  output$p_funding <- renderPlot({
    d <- county_data()
    req(nrow(d) > 0)
    ggplot(d, aes(x = reorder(fc_label, pct_funding),
                  y = pct_funding, fill = bar_color)) +
      geom_col(width = 0.6, alpha = 0.85) +
      geom_text(aes(label = paste0(round(pct_funding, 1), "%")),
                hjust = -0.2, color = "#0d1117", size = 5.5) +
      scale_fill_identity() +
      scale_y_continuous(limits = c(0, max(d$pct_funding) * 1.3)) +
      coord_flip() +
      labs(title = "Share of Safety Funding",
           x = NULL, y = "% of Total Safety Funding") +
      theme_odot()
  }, bg = "#0d1117")

  output$p_bubble <- renderPlot({
    d <- county_data()
    req(nrow(d) > 0)
    ggplot(d, aes(x = pct_funding, y = avg_crash_rate,
                  size = pct_fatalities, fill = bar_color)) +
      geom_point(shape = 21, alpha = 0.85,
                 color = "#0d1117", stroke = 1.5) +
      geom_text(aes(label = fc_label),
                vjust = -1.5, size = 5.0, color = "#0d1117") +
      scale_fill_identity() +
      scale_size_continuous(range = c(4, 14), name = "% Fatalities") +
      scale_x_continuous(labels = scales::percent_format(scale = 1)) +
      labs(title    = "Investment Gap",
           subtitle = "Bubble = share of fatalities",
           x = "% of Safety Funding",
           y = "Crash Rate (per MVMT)") +
      theme_odot() +
      theme(legend.position = "none")
  }, bg = "#0d1117")

  # ── Tab 2: Project Evaluator ─────────────────────────────────────────────

  # Pre-computed project data availability
  proj_available <- reactive({
    if (input$proj_scope == "By County") {
      project_segment_lookup %>%
        filter(grepl(input$proj_county, COUNTY_NME, 
                     ignore.case = TRUE))
    } else {
      project_segment_lookup
    }
  })

  proj_filtered <- reactive({
    p <- proj_available()
    if (!is.null(input$proj_category) && 
        input$proj_category != "All") {
      p <- p %>% 
        filter(PRIMARY_WORK_CATEGORY == input$proj_category)
    }
    p
  })

  output$proj_category_ui <- renderUI({
    base <- proj_available()

    counts <- base %>%
      group_by(PRIMARY_WORK_CATEGORY) %>%
      summarise(
        n_projects = n_distinct(PROJECT_NME),
        n_segments = n(),
        .groups    = "drop"
      ) %>%
      arrange(desc(n_projects)) %>%
      mutate(label = paste0(PRIMARY_WORK_CATEGORY,
                            " (", n_projects, " projects)"))

    total <- n_distinct(base$PROJECT_NME)

    choices <- c(
      setNames("All", paste0("All (", total, " projects)")),
      setNames(counts$PRIMARY_WORK_CATEGORY, counts$label)
    )

    selectInput("proj_category", "Project Type:",
                choices  = choices,
                selected = "All")
  })

  output$proj_name_ui <- renderUI({
    choices <- sort(unique(proj_filtered()$PROJECT_NME))
    if (length(choices) == 0) return(NULL)
    selectInput("proj_name", "Project:", choices = choices)
  })

  output$proj_details <- renderText({
    if (input$proj_view == "aggregate") return("")
    req(input$proj_name)
    p <- proj_filtered() %>%
      filter(PROJECT_NME == input$proj_name) %>%
      slice(1)
    if (nrow(p) == 0) return("")
    n_segs <- proj_filtered() %>%
      filter(PROJECT_NME == input$proj_name) %>%
      nrow()
    paste0(
      "Category:   ", p$PRIMARY_WORK_CATEGORY, "\n",
      "Completion: ", p$completion_year, "\n",
      "County:     ", p$COUNTY_NME, "\n",
      "Est. Cost:  $", scales::comma(p$EST_TOTAL_CONSTR_COST), "\n",
      "Segments:   ", n_segs, " matched"
    )
  })

  output$project_plot <- renderPlot({

    # ── AGGREGATE MODE ───────────────────────────────────────────────────
     if (input$proj_view == "aggregate") {
      req(input$proj_category)
      if (is.null(input$proj_category)) return(NULL)

     # If All selected use all segments, otherwise filter by type
      if (input$proj_category == "All") {
        base_data <- proj_available()
        type_label <- "All Project Types Combined"
      } else {
        base_data <- proj_available() %>%
          filter(PRIMARY_WORK_CATEGORY == input$proj_category)
        type_label <- input$proj_category
      }

       # base_data already set above

      if (nrow(base_data) == 0) {
        return(
          ggplot() +
            annotate("text", x = 0.5, y = 0.5,
                     label = paste0("No projects found for\n",
                                    input$proj_category,
                                    "\nin selected geography."),
                     color = COL_MUTED, size = 6,
                     hjust = 0.5, vjust = 0.5) +
            theme_void() +
            theme(plot.background = element_rect(fill = "#0d1117",
                                                 color = NA))
        )
      }

      agg_segs <- base_data %>%
        pull(ROADWAY_INVENTORY_ID) %>%
        unique()

      completion_lookup <- base_data %>%
        distinct(ROADWAY_INVENTORY_ID, completion_year)

      agg_data <- segment_year_app %>%
        filter(ROADWAY_INVENTORY_ID %in% agg_segs) %>%
        select(-any_of("completion_year")) %>%
        left_join(completion_lookup, by = "ROADWAY_INVENTORY_ID") %>%
        mutate(event_time = CRASH_YR - completion_year) %>%
        filter(event_time >= -2, event_time <= 2) %>%
        group_by(event_time) %>%
        summarise(
          avg_rate   = mean(crash_rate_MVMT, na.rm = TRUE),
          se         = sd(crash_rate_MVMT, na.rm = TRUE) / sqrt(n()),
          n_segments = n_distinct(ROADWAY_INVENTORY_ID),
          .groups    = "drop"
        )

      if (nrow(agg_data) == 0) {
        return(
          ggplot() +
            annotate("text", x = 0.5, y = 0.5,
                     label = "No data available for this project type.",
                     color = COL_MUTED, size = 8,
                     hjust = 0.5, vjust = 0.5) +
            theme_void() +
            theme(plot.background = element_rect(fill = "#0d1117",
                                                 color = NA))
        )
      }

      before_avg <- mean(agg_data$avg_rate[agg_data$event_time < 0],
                         na.rm = TRUE)
      after_avg  <- mean(agg_data$avg_rate[agg_data$event_time >= 0],
                         na.rm = TRUE)
      chg        <- round(after_avg - before_avg, 2)
      chg_label  <- ifelse(chg < 0,
                           paste0("▼ ", abs(chg), " per MVMT"),
                           paste0("▲ ", abs(chg), " per MVMT"))
      chg_color  <- case_when(
  chg < -0.5  ~ COL_GREEN,   # meaningful improvement
  chg >  0.5  ~ COL_RED,     # meaningful worsening
  TRUE        ~ COL_MUTED    # negligible change — gray
)
line_color <- chg_color
          n_proj <- n_distinct(base_data$PROJECT_NME)
      scope_label <- if (input$proj_scope == "Statewide") {
        "Ohio Statewide"
      } else {
        paste0(input$proj_county, " County")
      }

      ggplot(agg_data, aes(x = event_time, y = avg_rate)) +
        geom_ribbon(aes(ymin = avg_rate - 1.96 * se,
                        ymax = avg_rate + 1.96 * se),
                    fill = line_color, alpha = 0.15) +
        geom_line(color = line_color, linewidth = 1.3) +
        geom_point(color = line_color, size = 4) +
        geom_vline(xintercept = 0, linetype = "dashed",
                   color = COL_ORANGE, linewidth = 0.8) +
        annotate("text", x = 0.1,
                 y = max(agg_data$avg_rate +
                           1.96 * agg_data$se, na.rm = TRUE) * 0.97,
                 label = "Completion",
                 color = COL_ORANGE, hjust = 0, size = 5.5) +
        annotate("label",
                 x     = -1.5,
                 y     = max(agg_data$avg_rate +
                               1.96 * agg_data$se, na.rm = TRUE) * 0.6,
                 label = paste0(
                   "Before avg: ", round(before_avg, 1), "\n",
                   "After avg:  ", round(after_avg,  1), "\n",
                   "Change:     ", chg_label, "\n",
                   "Projects:   ", n_proj, "\n",
                   "Segments:   ", max(agg_data$n_segments)),
                 fill          = "#1c2128",
                 color         = chg_color,
                 size          = 5,
                 linewidth     = 0,
                 label.padding = unit(0.4, "lines"),
                 hjust         = 0) +
        scale_x_continuous(breaks = -2:2,
                           labels = c("-2", "-1", "0\n(Completion)",
                                      "+1", "+2")) +
        labs(
          title    = paste0(type_label, " — Aggregate Trend"),
        subtitle = paste0(scope_label,
                          " | 95% CI shown | n = ", n_proj,
                          " projects"),
          x        = "Years Relative to Completion",
          y        = "Avg Crash Rate (per MVMT)",
          caption  = "Source: ODOT Construction Projects & Crash Data 2020–2024"
        ) +
        theme_odot()

    # ── INDIVIDUAL MODE ──────────────────────────────────────────────────
    } else {
      req(input$proj_name)

      proj_segs <- proj_filtered() %>%
        filter(PROJECT_NME == input$proj_name)

      if (nrow(proj_segs) == 0) {
        return(
          ggplot() +
            annotate("text", x = 0.5, y = 0.5,
                     label = "No data available for this project.",
                     color = COL_MUTED, size = 8,
                     hjust = 0.5, vjust = 0.5) +
            theme_void() +
            theme(plot.background = element_rect(fill = "#0d1117",
                                                 color = NA))
        )
      }

      seg_ids   <- proj_segs$ROADWAY_INVENTORY_ID
      proj_info <- proj_segs %>% slice(1)
      comp_year <- proj_info$completion_year

      plot_data <- segment_year_app %>%
        filter(ROADWAY_INVENTORY_ID %in% seg_ids) %>%
        group_by(CRASH_YR) %>%
        summarise(
          avg_rate = mean(crash_rate_MVMT, na.rm = TRUE),
          n_segs   = n_distinct(ROADWAY_INVENTORY_ID),
          .groups  = "drop"
        )

      if (nrow(plot_data) == 0 || all(is.na(plot_data$avg_rate))) {
        return(
          ggplot() +
            annotate("text", x = 0.5, y = 0.5,
                     label = paste0("No crash data found for\n",
                                    input$proj_name),
                     color = COL_MUTED, size = 8,
                     hjust = 0.5, vjust = 0.5) +
            theme_void() +
            theme(plot.background = element_rect(fill = "#0d1117",
                                                 color = NA))
        )
      }

      avg_before <- mean(proj_segs$before_rate, na.rm = TRUE)
      avg_after  <- mean(proj_segs$after_rate,  na.rm = TRUE)
      chg        <- round(avg_after - avg_before, 2)
      chg_label  <- ifelse(chg < 0,
                           paste0("▼ ", abs(chg), " per MVMT"),
                           paste0("▲ ", abs(chg), " per MVMT"))
      chg_color  <- case_when(
  chg < -0.5  ~ COL_GREEN,   # meaningful improvement
  chg >  0.5  ~ COL_RED,     # meaningful worsening
  TRUE        ~ COL_MUTED    # negligible change — gray
)
line_color <- chg_color

      ggplot(plot_data, aes(x = CRASH_YR, y = avg_rate)) +
        geom_area(fill = line_color, alpha = 0.1) +
        geom_line(color = line_color, linewidth = 1.3) +
        geom_point(color = line_color, size = 4) +
        geom_vline(xintercept = comp_year, linetype = "dashed",
                   color = COL_ORANGE, linewidth = 0.8) +
        annotate("text",
                 x     = comp_year + 0.05,
                 y     = max(plot_data$avg_rate, na.rm = TRUE) * 0.95,
                 label = paste0("Completed ", comp_year),
                 color = COL_ORANGE, hjust = 0, size = 5.5) +
        annotate("label",
                 x     = mean(range(plot_data$CRASH_YR)),
                 y     = max(plot_data$avg_rate, na.rm = TRUE) * 0.5,
                 label = paste0(
                   "Before: ", round(avg_before, 1), "\n",
                   "After:  ", round(avg_after,  1), "\n",
                   "Change: ", chg_label, "\n",
                   "Segments: ", nrow(proj_segs)),
                 fill          = "#1c2128",
                 color         = chg_color,
                 size          = 5,
                 linewidth     = 0,
                 label.padding = unit(0.4, "lines")) +
        scale_x_continuous(breaks = unique(plot_data$CRASH_YR)) +
        labs(
          title    = input$proj_name,
          subtitle = paste0(proj_info$PRIMARY_WORK_CATEGORY,
                            " | ", proj_info$COUNTY_NME, " County",
                            " | Cost: $",
                            scales::comma(proj_info$EST_TOTAL_CONSTR_COST)),
          x        = "Year",
          y        = "Avg Crash Rate (per MVMT)"
        ) +
        theme_odot()
    }
  }, bg = "#0d1117")

  # ── Tab 3: Risk Explorer ─────────────────────────────────────────────────
  risk_data <- reactive({
    d <- analysis_app %>%
      filter(COUNTY_CD.x == input$risk_county,
             treated_safety == FALSE,
             !is.na(crashes_per_MVMT),
             ADT_TOTAL_NBR > 0)

    # Filter out very short segments if checkbox is checked
    if (input$risk_exclude_short) {
      d <- d %>%
        filter(SEGMENT_LENGTH_NBR >= input$risk_min_length)
    }

    if (input$risk_fc != "All") {
      d <- d %>%
        filter(FUNCTION_CLASS_CD == as.integer(input$risk_fc))
    }

    d %>%
      arrange(desc(crashes_per_MVMT)) %>%
      slice_head(n = input$risk_n) %>%
      mutate(
        fc_label      = fc_labels[as.character(FUNCTION_CLASS_CD)],
        crash_rate    = round(crashes_per_MVMT, 1),
        ADT_num       = ADT_TOTAL_NBR,  # keep numeric for sorting
        ADT           = scales::comma(ADT_TOTAL_NBR),
        length_mi     = round(SEGMENT_LENGTH_NBR, 3)
      ) %>%
      select(
        `Segment ID`  = ROADWAY_INVENTORY_ID,
        `Road/Route`  = NLF_ID,
        `Road Class`  = fc_label,
        `Crash Rate`  = crash_rate,
        `Fatalities`  = fatal_crashes,
        `ADT`         = ADT,
        `ADT_sort`    = ADT_num,       # hidden sort column
        `Length (mi)` = length_mi,
        `PCR Score`   = PCR_NBR
      )
  })

  output$risk_table <- renderDT({
    d <- risk_data()
    
    # Separate display and sort columns
    display_cols <- c("Segment ID", "Road/Route", "Road Class",
                      "Crash Rate", "Fatalities", "ADT",
                      "Length (mi)", "PCR Score")
    
    datatable(
      d,
      options = list(
        pageLength = 25,
        dom        = "frtip",
        scrollX    = TRUE,
        columnDefs = list(
          list(visible = FALSE, targets = 6),  # hide ADT_sort
          list(orderData = 6, targets = 5)     # sort ADT by ADT_sort
        )
      ),
      rownames  = FALSE,
      style     = "bootstrap",
      selection = "single",
      colnames  = c("Segment ID", "Road/Route", "Road Class",
                    "Crash Rate", "Fatalities", "ADT",
                    "ADT_sort", "Length (mi)", "PCR Score")
    ) %>%
      formatStyle(
        columns    = display_cols,
        color      = "#e6edf3",
        background = "#161b22"
      )
  })

  # Map under filters
  output$risk_map <- renderLeaflet({
    leaflet() %>%
      addProviderTiles(providers$CartoDB.DarkMatter) %>%
      setView(lng = -82.9, lat = 40.4, zoom = 9)
  })

  observeEvent(input$risk_table_rows_selected, {
    row <- input$risk_table_rows_selected
    req(row)

    selected_seg <- risk_data()[row, ]$`Segment ID`

    seg_geo <- segments_geo %>%
      filter(ROADWAY_INVENTORY_ID == selected_seg) %>%
      st_transform(4326)

    if (nrow(seg_geo) > 0) {
      center <- seg_geo %>%
        st_centroid() %>%
        st_coordinates()

      leafletProxy("risk_map") %>%
        clearShapes() %>%
        addPolylines(
          data   = seg_geo,
          color  = COL_RED,
          weight = 5,
          label  = paste0(
            "Segment: ", selected_seg, "\n",
            "Route: ", risk_data()[row, ]$`Road/Route`
          )
        ) %>%
        setView(lng = center[1], lat = center[2], zoom = 14)
    }
  })

  # ── Tab 4: Funding Gap ───────────────────────────────────────────────────
  output$funding_plot <- renderPlot({
    if (input$fund_scope == "By County") {
      d <- county_summary %>%
        filter(COUNTY_CD.x == input$fund_county)
    } else {
      d <- county_summary %>%
        group_by(FUNCTION_CLASS_CD) %>%
        summarise(
          n_segments       = sum(n_segments),
          n_treated        = sum(n_treated),
          total_crashes    = sum(total_crashes),
          total_fatalities = sum(total_fatalities),
          avg_crash_rate   = mean(avg_crash_rate, na.rm = TRUE),
          total_funding    = sum(total_funding),
          .groups          = "drop"
        ) %>%
        mutate(
          pct_funding    = round(100 * total_funding /
                                   sum(total_funding), 1),
          pct_fatalities = round(100 * total_fatalities /
                                   sum(total_fatalities), 1),
          pct_segments   = round(100 * n_treated / n_segments, 1),
          COUNTY_CD.x    = "Statewide"
        )
    }

    d <- d %>%
      mutate(
        fc_label  = fc_labels[as.character(FUNCTION_CLASS_CD)],
        bar_color = case_when(
          FUNCTION_CLASS_CD %in% c(1, 2) ~ COL_BLUE,
          FUNCTION_CLASS_CD == 3          ~ COL_YELLOW,
          FUNCTION_CLASS_CD %in% c(4, 5) ~ COL_RED,
          TRUE                            ~ COL_ORANGE
        ),
        x_var = if (input$fund_metric == "pct_funding")
          pct_funding else pct_segments
      ) %>%
      filter(!is.na(fc_label))

    ggplot(d, aes(x = x_var, y = avg_crash_rate,
                  size = pct_fatalities, fill = bar_color)) +
      geom_point(shape = 21, alpha = 0.85,
                 color = "#0d1117", stroke = 1.5) +
      geom_text(aes(label = fc_label),
                vjust = -1.5, size = 5.5, color = "#0d1117") +
      scale_fill_identity() +
      scale_size_continuous(range = c(5, 20),
                            name  = "% of Fatalities") +
      scale_x_continuous(
        labels = scales::percent_format(scale = 1)) +
      labs(
        title    = "Safety Investment Gap by Road Class",
        subtitle = "Bubble size = share of fatalities",
        x        = ifelse(input$fund_metric == "pct_funding",
                          "% of Total Safety Funding",
                          "% of Segments Receiving Investment"),
        y        = "Avg Crash Rate (per MVMT)",
        caption  = "Source: ODOT Construction Projects & Crash Data 2020–2024"
      ) +
      theme_odot()
  }, bg = "#0d1117")
  
  # ── Tab 5: Cost Effectiveness ────────────────────────────────────────────
  output$cost_plot <- renderPlot({

    # Build project-level summary from lookup
    cost_data <- project_segment_lookup %>%
      group_by(PROJECT_NME, PRIMARY_WORK_CATEGORY,
               COUNTY_NME, COUNTY_CD.x, completion_year,
               EST_TOTAL_CONSTR_COST) %>%
      summarise(
        avg_change  = mean(change, na.rm = TRUE),
        n_segments  = n(),
        avg_before  = mean(before_rate, na.rm = TRUE),
        avg_fc      = round(mean(FUNCTION_CLASS_CD, na.rm = TRUE)),
        .groups     = "drop"
      ) %>%
      filter(!is.na(EST_TOTAL_CONSTR_COST),
             EST_TOTAL_CONSTR_COST > 0,
             !is.na(avg_change))

    # Apply geography filter
    if (input$cost_scope == "By County") {
      cost_data <- cost_data %>%
        filter(grepl(input$cost_county, COUNTY_NME,
                     ignore.case = TRUE))
    }

    # Apply FC filter
    if (input$cost_fc != "All") {
      cost_data <- cost_data %>%
        filter(avg_fc == as.integer(input$cost_fc))
    }

    if (nrow(cost_data) == 0) {
      return(
        ggplot() +
          annotate("text", x = 0.5, y = 0.5,
                   label = "No projects found for selected filters.",
                   color = COL_MUTED, size = 8,
                   hjust = 0.5, vjust = 0.5) +
          theme_void() +
          theme(plot.background = element_rect(fill = "#0d1117",
                                               color = NA))
      )
    }

    # Color by outcome
    cost_data <- cost_data %>%
      mutate(
        outcome = case_when(
          avg_change < -0.5 ~ "Improved",
          avg_change >  0.5 ~ "Worsened",
          TRUE              ~ "No Change"
        ),
        dot_color = case_when(
          avg_change < -0.5 ~ COL_GREEN,
          avg_change >  0.5 ~ COL_RED,
          TRUE              ~ COL_MUTED
        ),
        cost_millions = EST_TOTAL_CONSTR_COST / 1e6
      )

    # Horizontal reference line
    ref_y <- 0

    # Label only the most extreme projects
    top_improved <- cost_data %>%
      slice_min(avg_change, n = 3)
    top_worsened <- cost_data %>%
      slice_max(avg_change, n = 3)
    top_costly   <- cost_data %>%
      slice_max(cost_millions, n = 2)

    label_data <- bind_rows(
      top_improved, top_worsened, top_costly
    ) %>% distinct(PROJECT_NME, .keep_all = TRUE)

    ggplot(cost_data, aes(x = cost_millions,
                          y = avg_change)) +
      # Reference line at zero
      geom_hline(yintercept = 0, linetype = "dashed",
                 color = COL_MUTED, linewidth = 0.6) +
      geom_hline(yintercept =  0.5, linetype = "dotted",
                 color = COL_RED, linewidth = 0.4, alpha = 0.5) +
      geom_hline(yintercept = -0.5, linetype = "dotted",
                 color = COL_GREEN, linewidth = 0.4, alpha = 0.5) +
      # Shaded regions
      annotate("rect",
               xmin = -Inf, xmax = Inf,
               ymin = -Inf, ymax = -0.5,
               fill = COL_GREEN, alpha = 0.04) +
      annotate("rect",
               xmin = -Inf, xmax = Inf,
               ymin = 0.5, ymax = Inf,
               fill = COL_RED, alpha = 0.04) +
      # Points
      geom_point(aes(size = n_segments, color = dot_color),
                 alpha = 0.75) +
      scale_color_identity() +
      # Labels for extreme projects
      geom_text(data = label_data,
                aes(label = PROJECT_NME),
                color    = "#0d1117",
                size     = 3.5,
                hjust    = -0.1,
                vjust    = 0.5,
                check_overlap = TRUE) +
      scale_size_continuous(range = c(3, 12),
                            name  = "Matched Segments") +
      scale_x_continuous(
        labels = scales::dollar_format(suffix = "M"),
        breaks = pretty(cost_data$cost_millions, n = 6)
      ) +
      # Region labels
      annotate("text",
               x     = max(cost_data$cost_millions) * 0.85,
               y     = min(cost_data$avg_change) * 0.85,
               label = "Effective",
               color = COL_GREEN, size = 5.5,
               fontface = "italic", alpha = 0.6) +
      annotate("text",
               x     = max(cost_data$cost_millions) * 0.85,
               y     = max(cost_data$avg_change) * 0.85,
               label = "Ineffective",
               color = COL_RED, size = 5.5,
               fontface = "italic", alpha = 0.6) +
      labs(
        title    = "Project Cost vs Safety Effectiveness",
        subtitle = paste0(
          if (input$cost_scope == "Statewide") "Ohio Statewide"
          else paste0(input$cost_county, " County"),
          " | Each point = one project | ",
          nrow(cost_data), " projects shown"
        ),
        x        = "Total Project Cost",
        y        = "Avg Crash Rate Change (per MVMT)",
        caption  = "Negative = crash rate decreased after completion | Source: ODOT 2020–2024"
      ) +
      theme_odot() +
      theme(legend.position = "right")
  }, bg = "#0d1117")

}

############################################
### RUN
############################################

shinyApp(ui = ui, server = server)
```
